---
title: "Meltwater source regions and the Glas δ¹⁸O record: importance, variability and alignment"
author: "Laura Endres"
date: today
format:
  docx:
    toc: false
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
set.seed(20260929)

dat <- function(f) read_csv(file.path("data", f), show_col_types = FALSE)

# Region codes used in the discharge files, in file order, mapped to the
# input-region names and surface volumes of Endres et al. (2026b, Tables 1-2).
regions <- tribble(
  ~code,      ~name,               ~ice,         ~V,
  "Med",      "EIS_MedSea",        "European",   1.79e13,
  "Bri",      "EIS_BayOfBiscay",   "European",   4.92e12,
  "Fen",      "EIS_NorwegianSea",  "European",   5.85e12,
  "EurArc",   "EIS_Arctic",        "European",   7.02e12,
  "AmeArc",   "LAU_Arctic",        "Laurentide", 1.34e13,
  "GIS",      "GIS_GreenlandSea",  "Greenland",  4.49e12,
  "NLau",     "LAU_LabradorSea",   "Laurentide", 1.33e13,
  "SLau",     "LAU_StLawrence",    "Laurentide", 1.58e13,
  "GulofMex", "LAU_GulfOfMexico",  "Laurentide", 1.19e13
)
d18O_ice <- -35   # end-member used for the main analysis

tidy_discharge <- function(d, tcol, qprefix, recon) {
  d <- d |> mutate(age = -.data[[tcol]])
  q <- d |> select(age, all_of(paste0(qprefix, regions$code))) |>
    rename_with(~ sub(paste0("^", qprefix), "", .x)) |>
    pivot_longer(-age, names_to = "code", values_to = "Q")
  a <- d |> select(age, ends_with(sprintf("(%.1f)", d18O_ice))) |>
    pivot_longer(-age, names_to = "col", values_to = "anom") |>
    mutate(code = sub(" d18O.*$", "", col)) |> select(-col)
  inner_join(q, a, by = c("age", "code")) |> mutate(recon = recon)
}
glac <- tidy_discharge(dat("Discharge_Glac1d_regional-withd18O.csv"), "time", "", "GLAC-1D")
ice6 <- tidy_discharge(dat("Discharge_ice6g_regional-withd18O.csv"), "t_adj", "discharge_", "ICE-6G")
long <- bind_rows(glac, ice6) |> left_join(regions, by = "code") |>
  mutate(name = factor(name, levels = regions$name),
         recon = factor(recon, levels = c("GLAC-1D", "ICE-6G")))
step <- c("GLAC-1D" = 100, "ICE-6G" = 500)

# Glas (entity 903) on its original chronology
entity <- dat("entity.csv") |> filter(entity_name == "Glas")
glas <- dat("sample.csv") |> filter(entity_id == entity$entity_id) |>
  inner_join(dat("original_chronology.csv"), by = "sample_id") |>
  inner_join(dat("d18O.csv"), by = "sample_id") |>
  arrange(interp_age)
win <- range(glas$interp_age)

# Scaling check: anomaly / (Q * d18O_ice) per region, times region volume
scal <- long |> filter(Q > 1e-6) |>
  group_by(recon, name, V) |>
  summarise(k = median(anom / (Q * d18O_ice)), .groups = "drop") |>
  mutate(kV = k * V)

fmt  <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
ka   <- function(x) sprintf("%.1f", x / 1000)
fmtp <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
```

# Questions

At the NISA location, which meltwater source region is the most important, and
which is the most variable, contributor to the δ¹⁸O anomaly over time? Is
there alignment between abrupt δ¹⁸O changes in the Glas data and the modelled
discharge, and how can that alignment be quantified?

# Data

**Discharge series.** `data/Discharge_Glac1d_regional-withd18O.csv` and
`data/Discharge_ice6g_regional-withd18O.csv` are the meltwater forcing series
of the impulse-response (Green's function) forward model of Endres et al.
(in prep.; `Impulse_Response_Revisions_v1.pdf` in this repository, Sect. 5.2.3
and 5.3.3). They give river-routed meltwater discharge (in Sv) into nine
North Atlantic dye-tracer source regions, computed from the GLAC-1D
reconstruction (`r fmt(max(glac$age)/1000, 0)`–`r fmt(abs(min(glac$age))/1000, 0)` ka,
`r step["GLAC-1D"]`-yr steps) and the ICE-6G reconstruction
(`r fmt(max(ice6$age)/1000, 0)` ka to `r fmt(abs(min(ice6$age))/1000, 0)` ka after
present, `r step["ICE-6G"]`-yr steps), following the meltwater protocol
described in the draft (after Ivanovic et al., 2018, and Romé et al., 2022,
as cited there). Time is taken as years BP
(negated `time` and `t_adj`); both series place their largest pulse at
`r ka(glac |> group_by(age) |> summarise(Q = sum(Q)) |> slice_max(Q, n = 1) |> pull(age))`
(GLAC-1D) and
`r ka(ice6 |> group_by(age) |> summarise(Q = sum(Q)) |> slice_max(Q, n = 1) |> pull(age))`
ka (ICE-6G), i.e. Meltwater Pulse 1A, which supports this convention. The
region codes are those of the forward-model draft (Med, Bri, Fen, EurArc,
AmeArc, GIS, NLau, SLau, GoM) and correspond, in the same order, to the
input regions of Endres et al. (2026b) (@tbl-regions). Each file also gives
the discharge converted to a δ¹⁸O anomaly per region for three ice-sheet
end-members (−30, −35, −40 ‰); this is the forcing that the forward model
convolves with the site kernels. Within every region the anomaly is a fixed
multiple of the discharge, and that multiple, times the region's surface
volume from Endres et al. (2026b), is the same constant for all nine regions
(`r fmt(mean(scal$kV)/1e13, 2)` ± `r fmt(sd(scal$kV)/1e13, 2)` × 10¹³ m³ s⁻¹ yr,
@tbl-regions), equal to one year of discharge. The anomaly columns are thus
the surface δ¹⁸O anomaly of the *source region* per year of injection,
Δδ¹⁸O = Q · 1 yr · δ¹⁸O~ice~ / V~region~, before propagation to NISA. The
convolved anomaly at the NISA site (Fig. 5.13C, E of the draft) is not part
of the repository data, so "contribution at NISA" is evaluated here on the
forcing side. All results below use the −35 ‰ end-member; the other
end-members rescale every anomaly by 6/7 or 8/7 and cannot change any
ranking, correlation or event timing.

**Glas.** Entity `r entity$entity_id` (La Vallina cave, NISA), `r nrow(glas)`
δ¹⁸O samples on the original Bchron chronology, `r ka(win[1])`–`r ka(win[2])`
ka BP (see the companion manuscript on the Glas age range). Abrupt δ¹⁸O
transitions identified by Endres et al. (2026a, Table 1) are used as
reference events.

```{r}
#| label: tbl-regions
#| tbl-cap: "Source-region codes in the discharge files, the corresponding input regions of Endres et al. (2026b), their surface volume V, and the empirical scaling k = anomaly / (Q · δ¹⁸O_ice) in the GLAC-1D file. k · V is constant, showing that the anomaly is the source-region anomaly per year of discharge."
scal |> filter(recon == "GLAC-1D") |>
  left_join(regions |> select(name, code, ice), by = "name") |>
  transmute(Code = code, `Input region` = as.character(name), `Ice sheet` = ice,
            `V (10¹² m³)` = fmt(V / 1e12, 2),
            `k (‰ Sv⁻¹ ‰⁻¹)` = fmt(k, 2),
            `k·V (10¹³)` = fmt(kV / 1e13, 2)) |>
  knitr::kable()
```

# Methods

**Importance and variability.** For each reconstruction the per-region
anomaly series was summarised over the full series and over the Glas
interval (`r ka(win[1])`–`r ka(win[2])` ka). *Importance* is the time-mean
anomaly (equivalently the time-integrated contribution) and its share of the
summed anomaly of all regions. *Variability* is the standard deviation of the
anomaly over time; the coefficient of variation (SD / |mean|) is given as a
scale-free complement. Because the anomaly is proportional to discharge, the
same rankings hold for discharge itself.

**Alignment with Glas.** Glas δ¹⁸O was averaged into the time bins of each
reconstruction (`r step["GLAC-1D"]` yr for GLAC-1D, `r step["ICE-6G"]` yr for
ICE-6G). Three complementary quantifications were used:

1. *Correlation at lag 0* between binned Glas δ¹⁸O and the total and
   per-region anomalies (Pearson r). A negative anomaly means fresher
   surface water, and a negative Glas δ¹⁸O shift is read as freshening, so
   alignment corresponds to positive r. Both series are strongly
   autocorrelated, so significance was assessed against 2,000 AR(1)
   surrogates of the Glas series with the same lag-1 autocorrelation.
2. *Lead–lag cross-correlation* between binned Glas δ¹⁸O and the GLAC-1D
   total anomaly for lags of ±2,000 yr, with the same surrogate envelope.
3. *Event matching.* Discharge pulses were defined on the GLAC-1D total
   anomaly as `r step["GLAC-1D"]`-yr steps whose change exceeds one standard
   deviation of all steps (freshening pulses: anomaly becoming more negative;
   recoveries: anomaly becoming less negative), merging consecutive steps
   into one pulse at the largest step. Glas events are the six δ¹⁸O
   transitions of Endres et al. (2026a, Table 1), plus data-driven steps
   where the change between consecutive `r step["GLAC-1D"]`-yr bins exceeds
   two standard deviations. A Glas event is matched if a discharge pulse of
   the same sign lies within ±300 yr. The number of matches was compared
   with 5,000 random placements of the same events in the Glas interval.

# Results

## Importance and variability of source regions

```{r}
#| label: summaries
summ <- function(d) d |>
  group_by(recon, ice, name) |>
  summarise(mean = mean(anom), sd = sd(anom), min = min(anom), .groups = "drop") |>
  group_by(recon) |>
  mutate(share = 100 * mean / sum(mean), cv = sd / abs(mean)) |>
  ungroup()
s_full <- summ(long)
s_win  <- summ(long |> filter(age >= win[1], age <= win[2]))
top <- function(s, r, what) s |> filter(recon == r) |>
  { \(x) if (what == "imp") slice_min(x, mean, n = 1) else slice_max(x, sd, n = 1) }()
imp_g <- top(s_win, "GLAC-1D", "imp"); imp_i <- top(s_win, "ICE-6G", "imp")
var_g <- top(s_win, "GLAC-1D", "var"); var_i <- top(s_win, "ICE-6G", "var")
impf_g <- top(s_full, "GLAC-1D", "imp"); impf_i <- top(s_full, "ICE-6G", "imp")
varf_g <- top(s_full, "GLAC-1D", "var"); varf_i <- top(s_full, "ICE-6G", "var")
ice_share <- s_win |> group_by(recon, ice) |> summarise(share = sum(share), .groups = "drop")
gs <- function(r, i) fmt(ice_share$share[ice_share$recon == r & ice_share$ice == i], 0)
```

@fig-stack shows the regional anomalies through time and @tbl-rank ranks the
regions within the Glas interval. The two reconstructions disagree on the
single most important region. In GLAC-1D the three largest contributors are
nearly tied: **`r imp_g$name`** (mean `r fmt(imp_g$mean)` ‰,
`r fmt(imp_g$share, 0)` % of the summed anomaly), followed by EIS_Arctic and
LAU_Arctic, and the European sectors together supply `r gs("GLAC-1D", "European")` %
of the anomaly versus `r gs("GLAC-1D", "Laurentide")` % from the Laurentide.
In ICE-6G the most important region is **`r imp_i$name`**
(mean `r fmt(imp_i$mean)` ‰, `r fmt(imp_i$share, 0)` %), and the Laurentide
sectors dominate (`r gs("ICE-6G", "Laurentide")` % versus
`r gs("ICE-6G", "European")` % European). Over the full series (@tbl-rankfull)
the most important region is `r impf_g$name` in GLAC-1D and
`r impf_i$name` in ICE-6G, because both reconstructions route most of
Meltwater Pulse 1A and the Holocene melt through the Laurentide sectors.

The most variable contribution over the Glas interval is
**`r var_g$name`** in GLAC-1D (SD `r fmt(var_g$sd)` ‰, with EIS_Arctic almost
equal) and **`r var_i$name`** in ICE-6G (SD `r fmt(var_i$sd)` ‰). In both
reconstructions the variability is concentrated in a few short pulses rather
than in a fluctuating background (@fig-stack): the regions with the largest SD
also have the most negative single values (@tbl-rank). GIS_GreenlandSea and
EIS_MedSea are negligible in GLAC-1D and small and steady in ICE-6G, so they
are the least variable in both cases.

```{r}
#| label: fig-stack
#| fig-cap: "Source-region δ¹⁸O anomalies (−35 ‰ end-member) through time for GLAC-1D (top) and ICE-6G (bottom), stacked so that the envelope is the summed anomaly of all nine regions. Grey shading marks the interval covered by Glas."
#| fig-width: 7
#| fig-height: 5.5
ggplot(long, aes(age / 1000, anom, fill = name)) +
  annotate("rect", xmin = win[1] / 1000, xmax = win[2] / 1000,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.5) +
  geom_area(position = "stack", colour = NA) +
  facet_wrap(~recon, ncol = 1, scales = "free_y") +
  scale_x_reverse(limits = c(28, -2)) +
  scale_fill_viridis_d(name = NULL, option = "turbo") +
  labs(x = "Age (ka BP)", y = expression(delta^18*O~anomaly~"(‰)")) +
  theme_bw(base_size = 10) + theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))
```

```{r}
#| label: tbl-rank
#| tbl-cap: "Source regions within the Glas interval, ranked by mean anomaly (importance). Share is the region's fraction of the summed anomaly; SD is the temporal standard deviation (variability) and CV = SD / |mean|. Values in ‰ for the −35 ‰ end-member."
rank_tab <- function(s) s |> arrange(recon, mean) |>
  transmute(Reconstruction = as.character(recon), Region = as.character(name),
            `Mean (‰)` = fmt(mean), `Share (%)` = fmt(share, 1),
            `Min (‰)` = fmt(min), `SD (‰)` = fmt(sd), CV = fmt(cv, 1)) |>
  mutate(Reconstruction = ifelse(duplicated(Reconstruction), "", Reconstruction))
knitr::kable(rank_tab(s_win))
```

```{r}
#| label: tbl-rankfull
#| tbl-cap: "As Table 2, but over the full length of each discharge series."
knitr::kable(rank_tab(s_full))
```

## Alignment between Glas δ¹⁸O and discharge

```{r}
#| label: align-prep
tot <- long |> group_by(recon, age) |>
  summarise(tot = sum(anom), Q = sum(Q), .groups = "drop")
bin_glas <- function(st) glas |>
  mutate(age = round(interp_age / st) * st) |>
  group_by(age) |> summarise(d18O = mean(d18O_measurement), n = n(), .groups = "drop")
joined <- lapply(names(step), function(r) {
  bin_glas(step[r]) |>
    inner_join(long |> filter(recon == r) |> select(age, name, anom), by = "age") |>
    inner_join(tot |> filter(recon == r) |> select(age, tot), by = "age") |>
    mutate(recon = r)
}) |> bind_rows()
wide <- joined |> select(recon, age, d18O, tot, name, anom) |>
  pivot_wider(names_from = name, values_from = anom) |>
  arrange(recon, desc(age))   # old -> young within each reconstruction

ar1_surrogates <- function(x, nsim = 2000) {
  n <- length(x); phi <- cor(x[-1], x[-n])
  m <- matrix(rnorm(n * nsim), n, nsim)
  for (i in 2:n) m[i, ] <- phi * m[i - 1, ] + m[i, ]
  list(phi = phi, sur = m)
}
lagcor <- function(x, y, k) {   # k > 0: y (discharge) leads x (Glas)
  n <- length(x)
  if (k >= 0) cor(x[(k + 1):n], y[1:(n - k)]) else cor(x[1:(n + k)], y[(1 - k):n])
}
cor_table <- function(r) {
  w <- wide |> filter(recon == r)
  S <- ar1_surrogates(w$d18O)
  cols <- c("tot", as.character(regions$name))
  tibble(series = cols) |> rowwise() |>
    mutate(r_obs = cor(w$d18O, w[[series]]),
           p = mean(abs(cor(S$sur, w[[series]])) >= abs(r_obs))) |>
    ungroup() |> mutate(recon = r, n = nrow(w), phi = S$phi)
}
ct <- bind_rows(cor_table("GLAC-1D"), cor_table("ICE-6G"))
ct_g <- ct |> filter(recon == "GLAC-1D"); ct_i <- ct |> filter(recon == "ICE-6G")
tot_g <- ct_g |> filter(series == "tot"); tot_i <- ct_i |> filter(series == "tot")
best_g <- ct_g |> filter(series != "tot") |> slice_max(r_obs, n = 1)
best_i <- ct_i |> filter(series != "tot") |> slice_max(r_obs, n = 1)
# first differences
wg <- wide |> filter(recon == "GLAC-1D")
r_diff <- cor(diff(wg$d18O), diff(wg$tot))
```

**Correlation at lag 0.** @tbl-cor lists the correlations. Binned Glas δ¹⁸O
is only weakly correlated with the GLAC-1D total anomaly
(r = `r fmt(tot_g$r_obs)`, n = `r tot_g$n` bins) and moderately with the
ICE-6G total (r = `r fmt(tot_i$r_obs)`, n = `r tot_i$n`). The strongest
regional correlation in both reconstructions is with
**`r best_g$series`** (r = `r fmt(best_g$r_obs)` in GLAC-1D,
`r fmt((ct_i |> filter(series == best_g$series))$r_obs)` in ICE-6G), the
sector that discharges into the Norwegian Sea close to the NISA moisture
source. EIS_BayOfBiscay is negatively correlated in GLAC-1D
(r = `r fmt((ct_g |> filter(series == "EIS_BayOfBiscay"))$r_obs)`) because its
discharge peaks at 22–23 ka while Glas records re-salinisation. The binned
Glas series has a lag-1 autocorrelation of `r fmt(tot_g$phi)` at 100-yr
resolution, so the effective sample size is small. Against AR(1) surrogates
only the EIS_NorwegianSea (positive) and EIS_BayOfBiscay (negative)
correlations in GLAC-1D reach p < 0.05
(p = `r fmtp((ct_g |> filter(series == "EIS_NorwegianSea"))$p)` and
`r fmtp((ct_g |> filter(series == "EIS_BayOfBiscay"))$p)`); the total
anomalies do not (p = `r fmtp(tot_g$p)` and `r fmtp(tot_i$p)`), and nothing
does in ICE-6G with `r tot_i$n` bins. Correlation of first differences is
likewise weak (r = `r fmt(r_diff)`). Level correlations therefore do not
demonstrate alignment: both series share a long-term deglacial trend and a
few pulses, which is not enough to separate from red noise.

```{r}
#| label: tbl-cor
#| tbl-cap: "Pearson correlation between binned Glas δ¹⁸O and the total and per-region source anomalies at lag 0. p is the two-sided fraction of AR(1) surrogates of the Glas series with |r| at least as large."
ct |> mutate(series = ifelse(series == "tot", "Total (all regions)", series)) |>
  select(series, recon, r_obs, p) |>
  pivot_wider(names_from = recon, values_from = c(r_obs, p)) |>
  transmute(Series = series,
            `r (GLAC-1D)` = fmt(`r_obs_GLAC-1D`), `p` = fmtp(`p_GLAC-1D`),
            `r (ICE-6G)` = fmt(`r_obs_ICE-6G`), `p ` = fmtp(`p_ICE-6G`)) |>
  knitr::kable()
```

```{r}
#| label: lags
lags <- seq(-20, 20)
S <- ar1_surrogates(wg$d18O)
cc <- tibble(lag = lags * step["GLAC-1D"],
             r = sapply(lags, function(k) lagcor(wg$d18O, wg$tot, k)),
             lo = sapply(lags, function(k) quantile(apply(S$sur, 2, lagcor, y = wg$tot, k = k), 0.025)),
             hi = sapply(lags, function(k) quantile(apply(S$sur, 2, lagcor, y = wg$tot, k = k), 0.975)))
cc_max <- cc |> slice_max(r, n = 1)
```

**Lead–lag.** The cross-correlation between Glas δ¹⁸O and the GLAC-1D total
anomaly (@fig-ccf) is highest at a lag of `r cc_max$lag` yr
(r = `r fmt(cc_max$r)`), i.e. with the Glas record **leading** the GLAC-1D
discharge by about `r abs(cc_max$lag)` yr, compared with
r = `r fmt(cc$r[cc$lag == 0])` at zero lag. The peak lies
`r ifelse(cc_max$r > cc_max$hi, "above", "within")` the 95 % surrogate
envelope. This is consistent with the observation in Endres et al. (2026a)
that GLAC-1D places the deglacial melt events roughly 500 yr later than Glas.

```{r}
#| label: fig-ccf
#| fig-cap: "Cross-correlation between binned Glas δ¹⁸O and the GLAC-1D total source anomaly. Positive lags mean the discharge leads Glas; negative lags mean Glas leads. Shading: 95 % envelope from 2,000 AR(1) surrogates of the Glas series."
#| fig-width: 6
#| fig-height: 3.2
ggplot(cc, aes(lag, r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey85") +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(colour = "#1f5f99") + geom_point(colour = "#1f5f99", size = 1.2) +
  labs(x = "Lag (yr; > 0 = discharge leads Glas)", y = "Pearson r") +
  theme_bw(base_size = 10)
```

```{r}
#| label: events
tg <- tot |> filter(recon == "GLAC-1D", age >= win[1] - 500, age <= win[2] + 500) |>
  arrange(desc(age)) |> mutate(dA = tot - lag(tot))
sd_dA <- sd(tg$dA, na.rm = TRUE)
pick_runs <- function(idx, val, keep_min) {
  if (!length(idx)) return(integer(0))
  runs <- split(idx, cumsum(c(1, diff(idx) != 1)))
  vapply(runs, function(ix) ix[if (keep_min) which.min(val[ix]) else which.max(val[ix])], integer(1))
}
detect_pulses <- function(mult) {
  thr <- mult * sd_dA
  i_f <- pick_runs(which(tg$dA < -thr), tg$dA, TRUE)
  i_r <- pick_runs(which(tg$dA >  thr), tg$dA, FALSE)
  bind_rows(tibble(idx = i_f, sign = -1), tibble(idx = i_r, sign = 1)) |>
    mutate(age = tg$age[idx], dA = tg$dA[idx]) |> arrange(desc(age))
}
thr <- 1 * sd_dA
pulses <- detect_pulses(1)

# Glas events: Table 1 of Endres et al. (2026a), d18O events only
ev_tab <- tribble(
  ~event,   ~mid,  ~lo,   ~hi,   ~sign,
  "23.4kS", 23440, 23330, 24050,  1,
  "17.8kS", 17840, 17820, 18030, -1,
  "16.4kS", 16440, 16140, 16550, -1,
  "16.1kS", 16070, 15900, 16380, -1,
  "15.3kS", 15270, 15140, 15360, -1,
  "14.7kS", 14680, 14540, 14810,  1
) |> mutate(source = "Endres et al. (2026a), Table 1")
# data-driven: 100-yr bin steps > 2 SD
wg2 <- wg |> mutate(dd = d18O - lag(d18O))
thr_g <- 2 * sd(wg2$dd, na.rm = TRUE)
i_g <- c(pick_runs(which(wg2$dd < -thr_g), wg2$dd, TRUE), pick_runs(which(wg2$dd > thr_g), wg2$dd, FALSE))
ev_dd <- tibble(event = sprintf("%.1fk (step)", wg2$age[i_g] / 1000), mid = wg2$age[i_g],
                lo = NA, hi = NA, sign = sign(wg2$dd[i_g]), source = "Bin step > 2 SD")
events <- bind_rows(ev_tab, ev_dd) |> arrange(desc(mid))

match_events <- function(ev, pl = pulses, tol = 300) {
  ev |> rowwise() |> mutate(
    pulse_age = { p <- pl$age[pl$sign == sign]; if (length(p)) p[which.min(abs(p - mid))] else NA_real_ },
    lead = mid - pulse_age,          # > 0: Glas event older, i.e. Glas leads
    matched = !is.na(lead) & abs(lead) <= tol) |> ungroup()
}
null_matches <- function(signs, pl = pulses, B = 5000, tol = 300) {
  replicate(B, {
    ages <- runif(length(signs), win[1], win[2])
    sum(mapply(function(a, s) { p <- pl$age[pl$sign == s]; length(p) && min(abs(p - a)) <= tol }, ages, signs))
  })
}
em <- match_events(events)
n_obs <- sum(em$matched[em$source != "Bin step > 2 SD"]); n_ev <- sum(em$source != "Bin step > 2 SD")
n_obs_all <- sum(em$matched); n_all <- nrow(em)
nm  <- null_matches(ev_tab$sign);  p_ev  <- mean(nm  >= n_obs)
nm2 <- null_matches(events$sign);  p_all <- mean(nm2 >= n_obs_all)
em_t1 <- em |> filter(source != "Bin step > 2 SD")
off_rng <- range(em_t1$lead); med_lead <- median(em_t1$lead)

# Sensitivity of the Table 1 event matching to pulse threshold and tolerance
sens <- expand_grid(mult = c(1, 1.5, 2), tol = c(200, 300, 500)) |> rowwise() |>
  mutate(pl = list(detect_pulses(mult)),
         n_pulses = nrow(pl),
         matched = sum(match_events(ev_tab, pl, tol)$matched),
         null = list(null_matches(ev_tab$sign, pl, B = 2000, tol = tol)),
         null_mean = mean(null), p = mean(null >= matched)) |>
  ungroup() |> select(-pl, -null)
# use the (1 SD, 300 yr) cell of the sensitivity grid as the headline null
p_ev <- (sens |> filter(mult == 1, tol == 300))$p
nm_mean <- (sens |> filter(mult == 1, tol == 300))$null_mean
```

**Event matching.** The GLAC-1D total anomaly within the Glas interval has
`r sum(pulses$sign == -1)` freshening pulses and `r sum(pulses$sign == 1)`
recoveries exceeding the 1-SD threshold of `r fmt(thr, 1)` ‰ per
`r step["GLAC-1D"]` yr (@fig-events, @tbl-events). Of the six δ¹⁸O events of
Endres et al. (2026a), `r n_obs` have a same-sign pulse within ±300 yr;
random placement of six such events gives `r fmt(nm_mean, 1)` matches on
average (p = `r fmtp(p_ev)`), because with `r nrow(pulses)` pulses in
`r fmt((win[2] - win[1]) / 1000, 1)` kyr a ±300-yr window is rarely empty.
Including the `r nrow(ev_dd)` data-driven steps, `r n_obs_all` of `r n_all`
events match (p = `r fmtp(p_all)`). @tbl-sens shows how this depends on the
pulse threshold and the tolerance. With tolerances of ±200–300 yr the match
count is at or above the random expectation for every threshold but not
significant (p ≥ `r fmtp(min(sens$p[sens$tol <= 300]))`). With a ±500-yr
tolerance `r (sens |> filter(mult == 1, tol == 500))$matched` of 6 events
are matched at the 1-SD threshold and
`r (sens |> filter(mult == 2, tol == 500))$matched` of 6 at the 2-SD
threshold (the `r (sens |> filter(mult == 2, tol == 500))$n_pulses` largest
pulses), with p = `r fmtp((sens |> filter(mult == 1, tol == 500))$p)` and
`r fmtp((sens |> filter(mult == 2, tol == 500))$p)`. The Glas events thus
coincide with GLAC-1D pulses at the ~500-yr level, which is also the scale of
the age-model offsets between the two archives, but not at the 100–300-yr
level. The offsets of the Table 1 events range from `r fmt(off_rng[1], 0)` to
`r fmt(off_rng[2], 0)` yr (median `r fmt(med_lead, 0)` yr, positive = Glas
earlier), i.e. the events show no consistent lead or lag at the event level;
the ~`r abs(cc_max$lag)`-yr lead in the cross-correlation reflects the broad
shape of the Heinrich Stadial 1 trough rather than the individual pulses.

The correspondences are specific rather than statistical. The Heinrich
Stadial 1 freshening events 16.4kS, 16.1kS and 15.3kS fall within
`r fmt(max(abs(em_t1$lead[em_t1$event %in% c("16.4kS", "16.1kS", "15.3kS")])), 0)` yr
of the GLAC-1D pulses at 16.2–16.4 and 15.2 ka. The abrupt freshening onset
at 17.8kS follows the GLAC-1D pulse at 18.2 ka (the Eurasian Arctic peak)
by `r fmt(-em_t1$lead[em_t1$event == "17.8kS"], 0)` yr, just outside the
±300-yr window; the forward-model draft discusses this either as a delayed
response to that Arctic pulse or, on the timing of AMOC proxies, as
meltwater retained at the surface by AMOC weakening. The re-salinisation at
14.7kS coincides with Meltwater Pulse 1A, the largest freshening pulse of the
series, an alignment of opposite sign that the draft reproduces as AMOC
re-strengthening cancelling the anomaly at NISA. The data-driven Glas steps
at 12.9–12.3 ka have no counterpart within 300 yr, and the large GLAC-1D
pulses of the Younger Dryas (13.3 and 11.7 ka) have no counterpart in Glas,
which the draft also notes as the main mismatch besides the LGM Bay of
Biscay discharge.

```{r}
#| label: tbl-sens
#| tbl-cap: "Sensitivity of the event matching (six Table 1 events) to the pulse threshold (multiples of the SD of 100-yr steps in the GLAC-1D total anomaly) and the matching tolerance. Null mean and p are from 2,000 random placements of the six events."
sens |> transmute(`Threshold (SD)` = mult, `Tolerance (yr)` = tol, `Pulses` = n_pulses,
                  `Matched (of 6)` = matched, `Null mean` = fmt(null_mean, 1), p = fmtp(p)) |>
  knitr::kable()
```

```{r}
#| label: tbl-events
#| tbl-cap: "Abrupt Glas δ¹⁸O events and the nearest same-sign pulse in the GLAC-1D total anomaly. Sign: −1 freshening, +1 re-salinisation. Offset is Glas age minus pulse age (positive: Glas earlier). Matched: |offset| ≤ 300 yr."
em |> transmute(Event = event, Source = source,
                `Midpoint (ka)` = ka(mid),
                `95 % CI (ka)` = ifelse(is.na(lo), "", paste0(ka(lo), "–", ka(hi))),
                Sign = sign, `Nearest pulse (ka)` = ifelse(is.na(pulse_age), "", ka(pulse_age)),
                `Offset (yr)` = ifelse(is.na(lead), "", fmt(lead, 0)),
                Matched = ifelse(matched, "yes", "no")) |>
  knitr::kable()
```

```{r}
#| label: fig-events
#| fig-cap: "Glas δ¹⁸O (top; points are samples, line is the 100-yr bin mean) and the total source anomaly of GLAC-1D (middle) and ICE-6G (bottom). Dashed lines: δ¹⁸O events of Endres et al. (2026a), blue = freshening, red = re-salinisation. Triangles: GLAC-1D pulses (down = freshening, up = recovery)."
#| fig-width: 7
#| fig-height: 6.5
xl <- c(win[2] / 1000 + 0.3, win[1] / 1000 - 0.3)
xbr <- seq(12, 24, 2)
evl <- geom_vline(data = ev_tab, aes(xintercept = mid / 1000, colour = factor(sign)),
                  linetype = "dashed", linewidth = 0.4)
colsc <- scale_colour_manual(values = c("-1" = "#1f5f99", "1" = "#b2182b"), guide = "none")
p1 <- ggplot() + evl +
  geom_point(data = glas, aes(interp_age / 1000, d18O_measurement), size = 0.5, colour = "grey55") +
  geom_line(data = wg, aes(age / 1000, d18O), colour = "black") +
  colsc + scale_x_reverse(limits = xl, breaks = xbr) +
  labs(x = NULL, y = expression(Glas~delta^18*O~"(‰)")) + theme_bw(base_size = 10)
p2 <- ggplot() + evl +
  geom_line(data = tot |> filter(recon == "GLAC-1D"), aes(age / 1000, tot), colour = "black") +
  geom_point(data = pulses, aes(age / 1000, dA * 0 + ifelse(sign < 0, -26, 1),
                                shape = factor(sign)), size = 2.2, colour = "black") +
  scale_shape_manual(values = c("-1" = 25, "1" = 24), guide = "none") +
  colsc + scale_x_reverse(limits = xl, breaks = xbr) +
  labs(x = NULL, y = "GLAC-1D anomaly (‰)") + theme_bw(base_size = 10)
p3 <- ggplot() + evl +
  geom_line(data = tot |> filter(recon == "ICE-6G"), aes(age / 1000, tot), colour = "black") +
  colsc + scale_x_reverse(limits = xl, breaks = xbr) +
  labs(x = "Age (ka BP)", y = "ICE-6G anomaly (‰)") + theme_bw(base_size = 10)
p1 / p2 / p3
```

# Answer

**Most important source region.** It depends on the ice-sheet reconstruction.
Within the Glas interval (`r ka(win[1])`–`r ka(win[2])` ka), GLAC-1D gives
`r imp_g$name` (`r fmt(imp_g$share, 0)` % of the summed source anomaly), with
EIS_Arctic and LAU_Arctic almost equal, and the European sectors together
provide `r gs("GLAC-1D", "European")` %. ICE-6G gives `r imp_i$name`
(`r fmt(imp_i$share, 0)` %), with the Laurentide sectors providing
`r gs("ICE-6G", "Laurentide")` %. Over the full series both reconstructions
put a Laurentide sector first (`r impf_g$name` and `r impf_i$name`).

**Most variable source region.** `r var_g$name` in GLAC-1D
(SD `r fmt(var_g$sd)` ‰ over the Glas interval; EIS_Arctic is a close second)
and `r var_i$name` in ICE-6G (SD `r fmt(var_i$sd)` ‰). In both cases the
variability comes from a small number of large, short pulses.

**Alignment with Glas.** Level correlations are weak to moderate
(r = `r fmt(tot_g$r_obs)` with GLAC-1D, `r fmt(tot_i$r_obs)` with ICE-6G) and
not significant against red-noise surrogates; the highest regional
correlation is with `r best_g$series`. Two quantifications do show
structure: (i) the lead–lag cross-correlation with GLAC-1D peaks with Glas
leading by about `r abs(cc_max$lag)` yr (r = `r fmt(cc_max$r)`, within the
red-noise envelope), and (ii) `r n_obs` of the six published abrupt Glas
events have a same-sign GLAC-1D pulse within ±300 yr, against
`r fmt(nm_mean, 1)` expected by chance (p = `r fmtp(p_ev)`), with offsets
between `r fmt(off_rng[1], 0)` and `r fmt(off_rng[2], 0)` yr; widening the
tolerance to ±500 yr matches `r (sens |> filter(mult == 2, tol == 500))$matched`
of six events against the `r (sens |> filter(mult == 2, tol == 500))$n_pulses`
largest pulses (p = `r fmtp((sens |> filter(mult == 2, tol == 500))$p)`), so
the coincidence is significant at the ~500-yr but not the centennial
scale. The alignment
is event-specific: the three Heinrich Stadial 1 freshening events match
GLAC-1D pulses to within a few hundred years, the 17.8 ka onset lags the
GLAC-1D Arctic pulse by ~`r fmt(-em_t1$lead[em_t1$event == "17.8kS"], 0)` yr,
and the 14.7 ka re-salinisation coincides with Meltwater Pulse 1A with
opposite sign. The appropriate quantification is therefore event-level
matching (offset per event, plus a Monte-Carlo null for the match count)
rather than correlation of the series, and, for a physically consistent
test, the comparison should be made with the convolved forward-model output
at NISA, which carries the AMOC-state dependence that explains the
opposite-sign case.

**Caveats.** The anomaly columns are the forward model's forcing: source-region
anomalies that scale with discharge divided by region volume. They do not
include the transport from each region to NISA, which Endres et al. (2026b)
and the forward-model draft show is weak for the Arctic sectors and strong
for the Bay of Biscay and Norwegian Sea sectors, nor the AMOC-mode dependence
of the kernels. Weighting by that transport would shift the "most important"
ranking further toward the European sectors. The event analysis uses the
GLAC-1D series only, because the 500-yr resolution of ICE-6G is coarser than
the events themselves.

# References

Endres, L., Pérez-Mejías, C., Ivanovic, R., Gregoire, L., Hughes, A. L. C.,
Cheng, H., & Stoll, H. (2026a). Interplay of North Atlantic freshening and deep
convection during the last deglaciation constrained by Iberian speleothems.
*Climate of the Past*, 22(4), 797–824. https://doi.org/10.5194/cp-22-797-2026

Endres, L., Ivanovic, R., Rome, Y., Tindall, J., Thurnherr, I., Sprenger, M.,
& Stoll, H. (2026b). Tracing meltwater from northern ice sheets to
palaeoclimate archives during the early last deglaciation: a conservative
tracer approach. *ESS Open Archive* (preprint).
https://doi.org/10.22541/essoar.15007256/v1

Endres, L., Ivanovic, R., Rome, Y., & Stoll, H. (in prep.). A dye tracer
forward-model for meltwater-related surface ocean anomalies. Draft
(`Impulse_Response_Revisions_v1.pdf`, this repository).