---
title: "Meltwater source regions and abrupt δ¹⁸O changes in stalagmite Glas: importance, variability and age-uncertain alignment"
author: "Laura Endres"
date: today
format:
  docx:
    toc: false
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(tidyr); library(purrr)
  library(ggplot2); library(patchwork); library(strucchange); library(Bchron)
})
dat <- function(f) read_csv(file.path("data", f), show_col_types = FALSE, name_repair = "unique_quiet")

fmt  <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
ka   <- function(x, d = 1) formatC(x / 1000, format = "f", digits = d)
fmtp <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
rng  <- function(x, d = 2) paste0(fmt(quantile(x, 0.05), d), " to ", fmt(quantile(x, 0.95), d))

# Region colours: the paper figures' colours (Endres et al. 2026b, Fig. 2a),
# lightness-tuned to pass colour-vision checks; identical to app/template.html.
reg_col <- c(EIS_MedSea = "#1f77b4", EIS_BayOfBiscay = "#f1780b", EIS_NorwegianSea = "#2ca02c",
             EIS_Arctic = "#d62728", LAU_Arctic = "#9467bd", GIS_GreenlandSea = "#9b4c3c",
             LAU_LabradorSea = "#de72bd", LAU_StLawrence = "#a5a608", LAU_GulfOfMexico = "#00b3c3")
# Stacking order (bottom to top) that keeps neighbouring colours distinguishable
stack_order <- c("EIS_BayOfBiscay", "LAU_LabradorSea", "LAU_StLawrence", "LAU_GulfOfMexico",
                 "EIS_Arctic", "EIS_MedSea", "EIS_NorwegianSea", "LAU_Arctic", "GIS_GreenlandSea")
col_fresh <- "#2a78d6"; col_resal <- "#eb6834"; ink2 <- "grey35"
theme_ms <- theme_bw(base_size = 9) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_line(linewidth = 0.2, colour = "grey90"),
        strip.background = element_blank(), strip.text = element_text(face = "bold", hjust = 0))
```

```{r}
#| label: data
regions <- tribble(
  ~code,      ~region,             ~ice,
  "Med",      "EIS_MedSea",        "European",
  "Bri",      "EIS_BayOfBiscay",   "European",
  "Fen",      "EIS_NorwegianSea",  "European",
  "EurArc",   "EIS_Arctic",        "European",
  "AmeArc",   "LAU_Arctic",        "Laurentide",
  "GIS",      "GIS_GreenlandSea",  "Greenland",
  "NLau",     "LAU_LabradorSea",   "Laurentide",
  "SLau",     "LAU_StLawrence",    "Laurentide",
  "GulofMex", "LAU_GulfOfMexico",  "Laurentide")
em <- -35   # ice-sheet end-member (per mil) used throughout

tidy_discharge <- function(d, tcol, qprefix, recon) {
  d <- d |> mutate(age = -.data[[tcol]])
  q <- d |> select(age, all_of(paste0(qprefix, regions$code))) |>
    rename_with(~ sub(paste0("^", qprefix), "", .x)) |>
    pivot_longer(-age, names_to = "code", values_to = "Q")
  a <- d |> select(age, ends_with(sprintf("(%.1f)", em))) |>
    pivot_longer(-age, names_to = "col", values_to = "anom") |>
    mutate(code = sub(" d18O.*$", "", col)) |> select(-col)
  inner_join(q, a, by = c("age", "code")) |> mutate(recon = recon)
}
long <- bind_rows(
  tidy_discharge(dat("Discharge_Glac1d_regional-withd18O.csv"), "time", "", "GLAC-1D"),
  tidy_discharge(dat("Discharge_ice6g_regional-withd18O.csv"), "t_adj", "discharge_", "ICE-6G")) |>
  left_join(regions, by = "code") |>
  mutate(region = factor(region, levels = regions$region),
         recon = factor(recon, levels = c("GLAC-1D", "ICE-6G")))
step <- c("GLAC-1D" = 100, "ICE-6G" = 500)

# Sanity checks: anomaly/discharge ratio constant within each region; MWP-1A timing
ratio_cv <- long |> filter(Q > 1e-5) |> group_by(recon, region) |>
  summarise(cv = sd(anom / Q) / abs(mean(anom / Q)), .groups = "drop")
tot <- long |> group_by(recon, age) |> summarise(tot = sum(anom), Q = sum(Q), .groups = "drop")
mwp <- tot |> group_by(recon) |> slice_max(Q, n = 1) |> ungroup()

# Glas (entity 903) on the published chronology, depth-ordered
entity <- dat("entity.csv") |> filter(entity_name == "Glas")
glas <- dat("sample.csv") |> filter(entity_id == entity$entity_id) |>
  inner_join(dat("original_chronology.csv"), by = "sample_id") |>
  inner_join(dat("d18O.csv"), by = "sample_id") |>
  arrange(depth_sample)
dates <- dat("dating.csv") |> filter(entity_id == entity$entity_id) |> arrange(depth_dating)
win <- range(glas$interp_age)
steps_g <- diff(sort(glas$interp_age))

# Transport weights from source region to proxy site (Endres et al. 2026b)
wts <- dat("meltmodel_site_weights.csv")
# Linear vs saturating conversion: for anomaly L (linear) and end-member em, the
# mixing-fraction anomaly is L / (1 + L / em), so the linear value is larger by |L| / |em|
max_sat <- max(abs(long$anom)) / abs(em)
```

```{r}
#| label: bchron
#| results: hide
# Age-model ensemble: Bchron on the U-Th dates, predicted at the sample depths
set.seed(20260929)
bch <- Bchronology(ages = dates$corr_age, ageSds = dates$corr_age_uncert_pos,
                   positions = dates$depth_dating, positionThicknesses = rep(0, nrow(dates)),
                   calCurves = rep("normal", nrow(dates)),
                   predictPositions = glas$depth_sample,
                   iterations = 12000, burn = 2000, thin = 10)
A <- bch$thetaPredict                           # N draws x 532 samples (depth order)
N <- nrow(A)
ens_med <- apply(A, 2, median)
ens_q   <- apply(A, 2, quantile, c(0.025, 0.05, 0.95, 0.975))
dev     <- ens_med - glas$interp_age
hw_ens  <- (ens_q["97.5%", ] - ens_q["2.5%", ]) / 2
hw_pub  <- (glas$interp_age_uncert_pos + glas$interp_age_uncert_neg) / 2
set.seed(20260930)
```

# Question

At the NISA location, which meltwater source region is the most important,
and which the most variable, contributor to the δ¹⁸O anomaly over time? Is
there alignment between abrupt δ¹⁸O changes in stalagmite Glas and the
modelled discharge, and how can that alignment be quantified?

# Data

**Discharge forcing.** `data/Discharge_Glac1d_regional-withd18O.csv` and
`data/Discharge_ice6g_regional-withd18O.csv` give river-routed meltwater
discharge (Sv) into nine North Atlantic dye-tracer source regions, derived
from the GLAC-1D (`r ka(max(long$age[long$recon == "GLAC-1D"]), 0)`–0 ka,
`r step[1]`-yr steps) and ICE-6G (`r ka(max(long$age[long$recon == "ICE-6G"]), 0)`
ka to `r ka(-min(long$age[long$recon == "ICE-6G"]), 0)` ka after present,
`r step[2]`-yr steps) ice-sheet reconstructions. They are the forcing of the
impulse-response (Green's function) forward model of Endres et al. (in prep.,
Sect. 5.2.3 and 5.3.5). Each file also gives the discharge converted to a
source-region δ¹⁸O anomaly for three ice-sheet end-members. Time is read as
years BP (negated `time` and `t_adj`). The total discharge peaks at
`r ka(mwp$age[1])` ka in GLAC-1D and `r ka(mwp$age[2])` ka in ICE-6G, the
Meltwater Pulse 1A interval, which confirms this convention. Within every
region the anomaly is a fixed multiple of the discharge (coefficient of
variation of the ratio ≤ `r fmt(100 * max(ratio_cv$cv), 1)` %), so the anomaly
is the discharge divided by the region's surface volume and scaled to the
end-member. The region codes map, in file order, to the input regions of
Endres et al. (2026b): EIS_MedSea, EIS_BayOfBiscay, EIS_NorwegianSea,
EIS_Arctic, LAU_Arctic, GIS_GreenlandSea, LAU_LabradorSea, LAU_StLawrence and
LAU_GulfOfMexico. All results use the −35 ‰ end-member. The −30 and −40 ‰
end-members rescale every anomaly by the same factor and change no ranking,
correlation or event timing.

The convolved anomaly at NISA (Endres et al., in prep., Fig. 5.13 C and E) is
not in the repository, and neither are the impulse-response kernels.
"Contribution at NISA" is therefore evaluated in two ways. On the *forcing
side*, the regions' source anomalies are compared before any ocean transport.
At the *proxy sites*, each source anomaly is weighted by the fraction of it
that reaches the site, using the transport weights of Endres et al. (2026b).

**Proxy sites and transport weights.** Endres et al. (2026b) evaluate ten
proxy sites. Nine of them have the information needed here (@tbl-sites):
six marine sediment cores, the NISA cave (La Vallina), Cave Without a Name
and the NGRIP ice core. Llarga Cave has no moisture-uptake mask and is also
absent from their Fig. 7. For each site, source region and AMOC mode (cold
and zonal), the paper gives the fraction of the source-region δ¹⁸O anomaly
that reaches the site after 500 yr of constant input. For land sites the
ocean anomaly is weighted by the site's moisture-uptake area. These weights
were taken from the paper's figure data
(`data/meltmodel_site_weights.csv`; provenance in
`data/meltmodel_site_weights.md`). They range from
`r fmt(min(wts$weight), 4)` to `r fmt(max(wts$weight), 3)`, so a site sees
at most about `r fmt(100 * max(wts$weight), 0)` % of a region's source anomaly.

**Glas.** Entity `r entity$entity_id` (La Vallina cave, NISA) has
`r nrow(glas)` δ¹⁸O samples between `r fmt(min(glas$depth_sample), 1)` and
`r fmt(max(glas$depth_sample), 1)` mm depth, dated to
`r ka(win[1])`–`r ka(win[2])` ka BP on the published Bchron chronology. The
median sampling step is `r fmt(median(steps_g), 0)` yr (mean
`r fmt(mean(steps_g), 0)` yr, maximum `r fmt(max(steps_g), 0)` yr), so Glas
is sampled several times per 100-yr GLAC-1D step. The age model rests on
`r nrow(dates)` U–Th dates (`r ka(min(dates$corr_age))`–`r ka(max(dates$corr_age))`
ka, uncertainties `r fmt(min(dates$corr_age_uncert_pos), 0)`–`r fmt(max(dates$corr_age_uncert_pos), 0)` yr).
The six abrupt δ¹⁸O transitions of Endres et al. (2026a, Table 1) are used as
published reference events, located by their sample-depth intervals.

# Methods

## Importance and variability of source regions

Each region's anomaly series was summarised over the Glas interval and over
the full series. To keep resolution from masquerading as variability, the
comparison between reconstructions uses a common 500-yr grid, with GLAC-1D
averaged into the ICE-6G time steps. GLAC-1D is also summarised at its native
100-yr resolution.

- **Importance** is the time-mean anomaly and the region's share of the summed
  anomaly of all regions. The *dominance fraction* is the share of time steps
  in which the region is the single largest contributor.
- **Variability** is the temporal standard deviation (SD) of the anomaly. The
  *variance share*, Cov(anomalyᵢ, total) / Var(total), apportions the
  variance of the total anomaly to the regions and sums to one. The SD of
  step-to-step changes measures abruptness.

**Anomaly at the proxy sites.** The anomaly at a site is approximated as the
weighted sum of the source anomalies, A~site~(t) = Σ~regions~ w~site,region~ ·
A~region~(t), separately for the cold and the zonal AMOC mode. This is the
equilibrium limit of the forward-model convolution. It applies the 500-yr
response to every time step at once and ignores transit times of years to
centuries. Slow changes are therefore represented about right, but short
pulses are overstated and appear without delay. The source anomalies are
those of the discharge files, which are linear in discharge and are what the
forward-model draft convolves. Endres et al. (2026b) instead use a mixing
fraction that saturates for large inputs, so for the largest pulse in these
series the linear anomaly is up to `r fmt(100 * max_sat, 0)` % larger. The
same importance and variability metrics as above are then computed for the
contributions at NISA.

## Age uncertainty of Glas

The published chronology is a single realisation. To propagate its
uncertainty, Bchron (Haslett and Parnell, 2008) was rerun on the `r nrow(dates)`
U–Th dates and evaluated at all sample depths, giving an ensemble of
`r fmt(N, 0)` monotonic age models. Following Rehfeld and Kurths (2014), every
alignment statistic below is computed for each ensemble member and reported
as the median and the 5–95 % range. Glas events are defined by sample depth,
not age, so each event is re-dated in every member. The discharge series have
their own chronological uncertainty, which is not given in the repository.
Its effect is bounded with a scan in which the discharge series is shifted by
−1,000 to +1,000 yr.

## Alignment of Glas with the discharge

The sign convention is that negative source anomalies and negative Glas δ¹⁸O
shifts both mean freshening, so alignment corresponds to positive correlation
and to same-sign events.

1. **Kernel cross-correlation.** Glas is irregularly sampled. Instead of
   interpolating it, correlations were estimated with the Gaussian-kernel
   estimator of Rehfeld et al. (2011) on the raw sample ages (bandwidth 25 yr;
   results change by < 0.01 between 15 and 100 yr). Both series were
   standardised within the Glas interval. The full series (*levels*) is
   dominated by millennial variations and the shared deglacial trend. A
   *high-pass* version isolates centennial variations: it is the residual
   after subtracting a Gaussian smoother of 750 yr width. Cross-correlations
   were computed for lags of ±2,000 yr against the total and the nine
   regional anomalies of GLAC-1D.

   *In plain terms:* the question is whether Glas δ¹⁸O is low when the
   modelled meltwater anomaly is large, and whether one follows the other
   with a delay. Glas samples are unevenly spaced, so instead of first
   interpolating Glas onto a regular grid, every pair of a Glas value and a
   discharge value contributes to the correlation with a weight that falls off
   the further apart the two are in time. This is done twice. The first pass
   uses the complete records, which mostly compares the slow deglacial trend.
   The second pass first subtracts a running average (a 750-yr Gaussian) from
   both records, so that only the fast, centennial wiggles remain, where
   abrupt events show up. Finally, one record is shifted against the other in
   100-yr steps, up to 2,000 yr in each direction, to see whether the
   agreement improves when Glas leads or lags the discharge.
2. **Significance.** Two nulls were used. The first uses 300 AR(1) surrogates
   of Glas on its irregular times, with persistence estimated from the data
   (Schulz and Mudelsee, 2002). The second uses 300 phase-randomised
   surrogates of the discharge that keep its spectrum (Theiler et al., 1992).
   For the lead–lag search, a *family-wise* p value compares the largest
   |r| over all lags, and over all ten series, with the same maximum in the
   surrogates. This accounts for testing many lags and regions (Franke and
   Donner, 2019).
3. **Event matching.** Discharge pulses are 100-yr changes in the GLAC-1D total
   anomaly larger than k SD of all changes (k = 1, 1.5 and 2), split into
   freshening pulses and recoveries; consecutive steps are merged. A Glas
   event is matched if a same-sign pulse lies within a tolerance of ±200, 300
   or 500 yr. For each event the ensemble gives a *match probability*, the
   fraction of age models in which it is matched. The number of matches is
   compared with 5,000 random placements of the same events in the Glas
   interval.
4. **Objective events.** To avoid relying only on published events,
   mean-shift breakpoints were fitted to the 100-yr binned Glas δ¹⁸O
   (Zeileis et al., 2002). Breakpoints were added only while the
   reduction in residual variance exceeded the 95th percentile of the same
   quantity in 200 AR(1) surrogates.
5. **Superposed epoch analysis.** Glas δ¹⁸O was composited in 100-yr bins from
   500 yr before to 1,000 yr after each GLAC-1D freshening pulse and each
   recovery, relative to the pre-pulse mean. The response statistic is the
   mean composite anomaly 100–600 yr after the pulse. The null places the same
   number of pulses at random within the Glas interval, 2,000 times.
6. **Detectability.** To judge what correlation *could* be expected, synthetic
   Glas series were built as a mix of the high-passed GLAC-1D total anomaly
   (signal) and AR(1) noise with the Glas persistence, for signal fractions of
   the variance between 5 and 50 %.
7. **ICE-6G.** At 500-yr resolution ICE-6G cannot resolve centennial events.
   It was compared with 500-yr Glas means through level and first-difference
   correlations only.

# Results

## Importance and variability of source regions

```{r}
#| label: part-a
g500 <- long |> filter(recon == "GLAC-1D") |> mutate(age = round(age / 500) * 500) |>
  group_by(recon, age, code, region, ice) |> summarise(Q = mean(Q), anom = mean(anom), .groups = "drop")
c500 <- bind_rows(g500, long |> filter(recon == "ICE-6G"))
summ <- function(d) d |>
  group_by(recon, age) |> mutate(tot = sum(anom), dom = anom == min(anom)) |> ungroup() |>
  arrange(recon, region, desc(age)) |> group_by(recon, region) |> mutate(dA = anom - lag(anom)) |> ungroup() |>
  group_by(recon, ice, region) |>
  summarise(mean = mean(anom), sd = sd(anom), min = min(anom), dom = mean(dom),
            vshare = cov(anom, tot) / var(tot), sd_diff = sd(dA, na.rm = TRUE), .groups = "drop") |>
  group_by(recon) |> mutate(share = mean / sum(mean)) |> ungroup()
s_win    <- summ(c500 |> filter(age >= win[1], age <= win[2]))
s_full   <- summ(c500)
s_win100 <- summ(long |> filter(recon == "GLAC-1D", age >= win[1], age <= win[2]))
pick <- function(s, r, v, f = which.max) { x <- s |> filter(recon == r); x[f(x[[v]]), ] }
imp  <- function(s, r) pick(s, r, "share")
var_ <- function(s, r) pick(s, r, "sd")
vsh  <- function(s, r) pick(s, r, "vshare")
top3 <- s_win |> filter(recon == "GLAC-1D") |> slice_max(share, n = 3)
ice_share <- s_win |> group_by(recon, ice) |> summarise(share = sum(share), .groups = "drop")
gs <- function(r, i) fmt(100 * ice_share$share[ice_share$recon == r & ice_share$ice == i], 0)
```

Within the Glas interval (@tbl-rank, @fig-stack), the two reconstructions
disagree on the most important region.

- **GLAC-1D.** Three regions are almost tied: `r top3$region[1]`
  (`r fmt(100 * top3$share[1], 1)` % of the summed anomaly),
  `r top3$region[2]` (`r fmt(100 * top3$share[2], 1)` %) and `r top3$region[3]`
  (`r fmt(100 * top3$share[3], 1)` %). European sectors supply
  `r gs("GLAC-1D", "European")` % and Laurentide sectors
  `r gs("GLAC-1D", "Laurentide")` %. `r pick(s_win, "GLAC-1D", "dom")$region`
  is the largest single contributor most often
  (`r fmt(100 * pick(s_win, "GLAC-1D", "dom")$dom, 0)` % of time steps),
  because it dominates the Last Glacial Maximum while contributing little
  afterwards.
- **ICE-6G.** `r imp(s_win, "ICE-6G")$region` is the most important region
  (`r fmt(100 * imp(s_win, "ICE-6G")$share, 1)` %), and Laurentide sectors
  supply `r gs("ICE-6G", "Laurentide")` %.

The most variable region is `r var_(s_win, "GLAC-1D")$region` in GLAC-1D
(SD `r fmt(var_(s_win, "GLAC-1D")$sd)` ‰, variance share
`r fmt(100 * vsh(s_win, "GLAC-1D")$vshare, 0)` %). In ICE-6G the largest SD
belongs to `r var_(s_win, "ICE-6G")$region`
(`r fmt(var_(s_win, "ICE-6G")$sd)` ‰), and `r vsh(s_win, "ICE-6G")$region`
carries the largest variance share
(`r fmt(100 * vsh(s_win, "ICE-6G")$vshare, 0)` %). At native 100-yr
resolution, GLAC-1D gives the same answer: `r var_(s_win100, "GLAC-1D")$region`
has the largest SD (`r fmt(var_(s_win100, "GLAC-1D")$sd)` ‰) and the largest
variance share (`r fmt(100 * vsh(s_win100, "GLAC-1D")$vshare, 0)` %). The
variability is carried by a few short pulses, so the most variable regions
also reach the most negative single values. GIS_GreenlandSea and EIS_MedSea
are negligible in GLAC-1D (< `r fmt(100 * max(s_win$share[s_win$recon == "GLAC-1D" & s_win$region %in% c("GIS_GreenlandSea", "EIS_MedSea")]), 1)` % each). In ICE-6G they
contribute `r fmt(100 * min(s_win$share[s_win$recon == "ICE-6G" & s_win$region %in% c("GIS_GreenlandSea", "EIS_MedSea")]), 1)`–`r fmt(100 * max(s_win$share[s_win$recon == "ICE-6G" & s_win$region %in% c("GIS_GreenlandSea", "EIS_MedSea")]), 1)` %
each, about as much as EIS_BayOfBiscay, but they vary little.

Over the full series (@tbl-rankfull), `r imp(s_full, "GLAC-1D")$region` is the
most important region in both GLAC-1D and ICE-6G, because both reconstructions
route most of Meltwater Pulse 1A and the Holocene melt through Laurentide
sectors.

@fig-importance shows these shares directly. Within the Glas interval,
GLAC-1D spreads the anomaly over many regions with a European majority, while
ICE-6G concentrates it in the Laurentide sectors. Over the full series both
reconstructions become Laurentide-dominated. @fig-radar compares all five
metrics of @tbl-rank and @tbl-rankfull region by region.

```{r}
#| label: fig-importance
#| fig-cap: "Importance of the source regions: share of the summed source δ¹⁸O anomaly (500-yr grid, −35 ‰ end-member) within the Glas interval and over the full series. Regions are grouped by ice sheet (European, Greenland, Laurentide from left); labels give shares of at least 7 %, and the text at the right gives the European and Laurentide totals."
#| fig-width: 7
#| fig-height: 3.4
ice_order <- c("EIS_MedSea", "EIS_BayOfBiscay", "EIS_NorwegianSea", "EIS_Arctic", "GIS_GreenlandSea",
               "LAU_Arctic", "LAU_LabradorSea", "LAU_StLawrence", "LAU_GulfOfMexico")
imp_df <- bind_rows(s_win |> mutate(period = "Glas interval"), s_full |> mutate(period = "Full series")) |>
  mutate(bar = factor(paste(recon, period, sep = " · "),
                      levels = rev(c("GLAC-1D · Glas interval", "ICE-6G · Glas interval",
                                     "GLAC-1D · Full series", "ICE-6G · Full series"))),
         region = factor(as.character(region), levels = rev(ice_order))) |>
  arrange(bar, desc(region)) |> group_by(bar) |>
  mutate(xmax = cumsum(share), xmin = xmax - share, xmid = (xmin + xmax) / 2) |> ungroup()
imp_ice <- imp_df |> group_by(bar, ice) |> summarise(share = sum(share), .groups = "drop") |>
  pivot_wider(names_from = ice, values_from = share) |>
  mutate(lab = sprintf("Eur %s %% · Lau %s %%", fmt(100 * European, 0), fmt(100 * Laurentide, 0)))
ggplot(imp_df) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = as.numeric(bar) - 0.38, ymax = as.numeric(bar) + 0.38, fill = region),
            colour = "white", linewidth = 0.4) +
  geom_text(data = imp_df |> filter(share >= 0.07), aes(xmid, as.numeric(bar), label = fmt(100 * share, 0)),
            size = 2.6, colour = "white", fontface = "bold") +
  geom_text(data = imp_ice, aes(1.02, as.numeric(bar), label = lab), hjust = 0, size = 2.6, colour = "grey25") +
  scale_y_continuous(breaks = seq_along(levels(imp_df$bar)), labels = levels(imp_df$bar), expand = expansion(add = 0.3)) +
  scale_x_continuous(labels = function(x) paste0(100 * x, " %"), breaks = seq(0, 1, 0.25), expand = expansion(mult = c(0, 0))) +
  coord_cartesian(xlim = c(0, 1.33), clip = "off") +
  scale_fill_manual(values = reg_col, breaks = ice_order, name = NULL) +
  labs(x = "Share of summed source anomaly", y = NULL) +
  theme_ms + theme(legend.position = "bottom", panel.grid.major.y = element_blank(),
                   plot.margin = margin(5, 10, 5, 5)) +
  guides(fill = guide_legend(nrow = 3))
```

```{r}
#| label: fig-radar
#| fig-cap: "Region profiles for the five metrics of Tables 1 and 2 (500-yr grid). Each spoke is scaled to the largest value of that metric over all regions, both reconstructions and both periods, so the outer ring is the maximum. Solid lines: Glas interval; dashed: full series. Negative variance shares are drawn at zero."
#| fig-width: 7
#| fig-height: 7
metrics <- c(share = "Share", dom = "Dominant", sd = "SD", vshare = "Var. share", sd_diff = "SD Δ")
rad <- bind_rows(s_win |> mutate(period = "Glas interval"), s_full |> mutate(period = "Full series")) |>
  select(recon, period, region, all_of(names(metrics))) |>
  pivot_longer(all_of(names(metrics)), names_to = "metric", values_to = "value") |>
  mutate(value = pmax(value, 0)) |>
  group_by(metric) |> mutate(r = value / max(value)) |> ungroup() |>
  mutate(k = match(metric, names(metrics)), ang = pi / 2 - 2 * pi * (k - 1) / length(metrics),
         x = r * cos(ang), y = r * sin(ang),
         region = factor(as.character(region), levels = as.character(regions$region)),
         period = factor(period, levels = c("Glas interval", "Full series"))) |>
  arrange(recon, period, region, k)
rad_closed <- rad |> group_by(recon, period, region) |> group_modify(~ bind_rows(.x, .x[1, ])) |> ungroup()
spokes <- tibble(k = seq_along(metrics), ang = pi / 2 - 2 * pi * (k - 1) / length(metrics),
                 x = cos(ang), y = sin(ang), lab = unname(metrics),
                 lx = 1.28 * cos(ang), ly = 1.2 * sin(ang))
rings <- expand_grid(r = c(0.5, 1), t = seq(0, 2 * pi, length.out = 73)) |> mutate(x = r * cos(t), y = r * sin(t))
rec_col <- c("GLAC-1D" = "#184f95", "ICE-6G" = "#c26a00")
ggplot() +
  geom_path(data = rings, aes(x, y, group = r), colour = "grey85", linewidth = 0.3) +
  geom_segment(data = spokes, aes(0, 0, xend = x, yend = y), colour = "grey85", linewidth = 0.3) +
  geom_text(data = spokes, aes(lx, ly, label = lab), size = 2.3, colour = "grey35") +
  geom_path(data = rad_closed, aes(x, y, colour = recon, linetype = period, group = interaction(recon, period)), linewidth = 0.6) +
  facet_wrap(~region, ncol = 3) +
  scale_colour_manual(values = rec_col, name = NULL) +
  scale_linetype_manual(values = c("Glas interval" = "solid", "Full series" = "22"), name = NULL) +
  coord_equal(xlim = c(-1.45, 1.45), ylim = c(-1.2, 1.3)) +
  theme_void(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8.5, margin = margin(b = 2)), legend.position = "bottom",
        panel.spacing = unit(4, "pt"))
```

```{r}
#| label: fig-stack
#| fig-cap: "Source-region δ¹⁸O anomalies (−35 ‰ end-member) at native resolution, stacked so the lower envelope is the summed anomaly. Colours follow the region colours of Endres et al. (2026b). Grey shading marks the Glas interval."
#| fig-width: 7
#| fig-height: 5
# geom_area stacks the first factor level on top, so reverse the bottom-to-top order
stk <- long |> mutate(grp = factor(as.character(region), levels = rev(stack_order)))
ggplot(stk, aes(age / 1000, anom, fill = grp)) +
  annotate("rect", xmin = win[1] / 1000, xmax = win[2] / 1000, ymin = -Inf, ymax = Inf, fill = "grey92") +
  geom_area(position = "stack", colour = NA) +
  facet_wrap(~recon, ncol = 1, scales = "free_y") +
  scale_x_reverse(limits = c(28, -2), breaks = seq(28, 0, -4)) +
  scale_fill_manual(values = reg_col, breaks = as.character(regions$region), name = NULL) +
  labs(x = "Age (ka BP)", y = expression("Source" ~ delta^18*O ~ "anomaly (‰)")) +
  theme_ms + theme(legend.position = "bottom") + guides(fill = guide_legend(nrow = 3))
```

```{r}
#| label: tbl-rank
#| tbl-cap: "Source regions within the Glas interval on the common 500-yr grid, ranked by importance. Share: fraction of the summed anomaly. Dominant: fraction of time steps in which the region is the largest contributor. SD: temporal standard deviation. Var. share: Cov(region, total)/Var(total). SD Δ: standard deviation of 500-yr changes."
rank_tab <- function(s) s |> arrange(recon, desc(share)) |>
  transmute(Reconstruction = as.character(recon), Region = as.character(region),
            `Mean (‰)` = fmt(mean), `Share (%)` = fmt(100 * share, 1), `Dominant (%)` = fmt(100 * dom, 0),
            `SD (‰)` = fmt(sd), `Var. share (%)` = fmt(100 * vshare, 0), `SD Δ (‰)` = fmt(sd_diff)) |>
  mutate(Reconstruction = ifelse(duplicated(Reconstruction), "", Reconstruction))
knitr::kable(rank_tab(s_win))
```

```{r}
#| label: tbl-rankfull
#| tbl-cap: "As the previous table, but over the full length of each series."
knitr::kable(rank_tab(s_full))
```

## Transport-weighted anomaly at the proxy sites

```{r}
#| label: sites
site_ts <- long |> select(recon, age, region, anom) |> mutate(region = as.character(region)) |>
  inner_join(wts |> select(site, amoc_mode, region, weight), by = "region", relationship = "many-to-many") |>
  group_by(recon, amoc_mode, site, age) |> summarise(anom = sum(weight * anom), .groups = "drop")
site_info <- wts |> distinct(site, type, lat, lon) |> arrange(desc(lat))
site_sum <- site_ts |> filter(age >= win[1], age <= win[2]) |>
  group_by(recon, amoc_mode, site) |> summarise(mean = mean(anom), min = min(anom), .groups = "drop")
nisa_c <- long |> filter(age >= win[1], age <= win[2]) |> select(recon, age, region, anom) |>
  mutate(region = as.character(region)) |>
  inner_join(wts |> filter(site == "La Vallina (NISA)") |> select(amoc_mode, region, weight),
             by = "region", relationship = "many-to-many") |>
  mutate(c = weight * anom) |>
  group_by(recon, amoc_mode, region) |> summarise(mean = mean(c), sd = sd(c), .groups = "drop") |>
  group_by(recon, amoc_mode) |> mutate(share = mean / sum(mean)) |> ungroup()
nisa_top <- nisa_c |> group_by(recon, amoc_mode) |> slice_max(share, n = 1) |> ungroup()
nisa_var <- nisa_c |> group_by(recon, amoc_mode) |> slice_max(sd, n = 1) |> ungroup()
nt <- function(r, m, what = nisa_top) what |> filter(recon == r, amoc_mode == m)
ss <- function(r, m, s, v = "mean") site_sum[[v]][site_sum$recon == r & site_sum$amoc_mode == m & site_sum$site == s]
zc <- site_sum |> select(recon, amoc_mode, site, mean) |> pivot_wider(names_from = amoc_mode, values_from = mean) |>
  mutate(ratio = zonal / cold)
var_same <- nisa_var |> group_by(recon) |> summarise(same = n_distinct(region) == 1) |> pull(same) |> all()
```

@fig-sites shows the transport-weighted anomaly at all nine sites. The
anomaly is largest in the Nordic Seas and the subpolar North Atlantic and
smallest at the continental sites. PS2644-5 averages
`r fmt(ss("GLAC-1D", "cold", "PS2644-5"))` ‰ over the Glas interval with
GLAC-1D in the cold mode, compared with
`r fmt(ss("GLAC-1D", "cold", "La Vallina (NISA)"))` ‰ at NISA. In the zonal
mode, with a strong AMOC, the sites receive between
`r fmt(100 * min(zc$ratio), 0)` and `r fmt(100 * max(zc$ratio), 0)` % of the
cold-mode anomaly (median `r fmt(100 * median(zc$ratio), 0)` %), because more
of the meltwater is exported from the surface North Atlantic.
@tbl-sites lists the sites and their Glas-interval means.

At NISA, weighting by transport changes the ranking of the source regions
(@fig-nisa). In GLAC-1D, `r nt("GLAC-1D", "cold")$region` becomes the most
important contributor, with `r fmt(100 * nt("GLAC-1D", "cold")$share, 0)` % of
the anomaly in the cold mode and `r fmt(100 * nt("GLAC-1D", "zonal")$share, 0)` %
in the zonal mode. The Bay of Biscay transports efficiently to the cave's
moisture source, whereas the two Arctic sectors, which rank near the top on
the source side, reach NISA only weakly. In ICE-6G, `r nt("ICE-6G", "cold")$region`
stays first, with `r fmt(100 * nt("ICE-6G", "cold")$share, 0)` % (cold) and
`r fmt(100 * nt("ICE-6G", "zonal")$share, 0)` % (zonal). The most variable
contribution at NISA is `r nt("GLAC-1D", "cold", nisa_var)$region` in GLAC-1D
and `r nt("ICE-6G", "cold", nisa_var)$region` in ICE-6G
`r ifelse(var_same, "in both AMOC modes", "in the cold AMOC mode")`.

```{r}
#| label: fig-sites
#| fig-cap: "Transport-weighted δ¹⁸O anomaly at the nine proxy sites of Endres et al. (2026b), ordered from north to south: the sum over regions of the source anomaly (−35 ‰ end-member) times the site's transport weight. Solid lines: cold (weak) AMOC mode; dashed: zonal (strong) mode. Grey shading marks the Glas interval. This is an equilibrium approximation that ignores transit times, so short pulses are overstated."
#| fig-width: 7
#| fig-height: 6.2
rec_col <- c("GLAC-1D" = "#184f95", "ICE-6G" = "#c26a00")
ggplot(site_ts |> mutate(site = factor(site, levels = site_info$site)),
       aes(age / 1000, anom, colour = recon, linetype = amoc_mode)) +
  annotate("rect", xmin = win[1] / 1000, xmax = win[2] / 1000, ymin = -Inf, ymax = Inf, fill = "grey92") +
  geom_hline(yintercept = 0, linewidth = 0.25, colour = "grey60") +
  geom_line(linewidth = 0.35) +
  facet_wrap(~site, ncol = 3) +
  scale_x_reverse(limits = c(28, 0), breaks = seq(28, 0, -7)) +
  scale_colour_manual(values = rec_col, name = NULL) +
  scale_linetype_manual(values = c(cold = "solid", zonal = "22"), labels = c(cold = "Cold AMOC mode", zonal = "Zonal AMOC mode"), name = NULL) +
  labs(x = "Age (ka BP)", y = expression("Site" ~ delta^18*O ~ "anomaly (‰)")) +
  theme_ms + theme(legend.position = "bottom")
```

```{r}
#| label: tbl-sites
#| tbl-cap: "Proxy sites of Endres et al. (2026b) used here, from north to south, with the mean transport-weighted δ¹⁸O anomaly over the Glas interval (‰, −35 ‰ end-member) for each reconstruction and AMOC mode."
site_info |>
  left_join(site_sum |> mutate(col = paste(recon, amoc_mode)) |> select(site, col, mean) |>
              pivot_wider(names_from = col, values_from = mean), by = "site") |>
  transmute(Site = site, Type = type, `Lat (°N)` = fmt(lat, 1), `Lon (°E)` = fmt(lon, 1),
            `GLAC-1D cold` = fmt(`GLAC-1D cold`), `GLAC-1D zonal` = fmt(`GLAC-1D zonal`),
            `ICE-6G cold` = fmt(`ICE-6G cold`), `ICE-6G zonal` = fmt(`ICE-6G zonal`)) |>
  knitr::kable()
```

```{r}
#| label: fig-nisa
#| fig-cap: "Importance of the source regions at NISA after weighting by transport: share of the transport-weighted anomaly within the Glas interval, for each reconstruction and AMOC mode. Compare with the source-side shares in @fig-importance; labels give shares of at least 7 %."
#| fig-width: 7
#| fig-height: 3
nb <- nisa_c |>
  mutate(bar = factor(paste(recon, amoc_mode, sep = " · "),
                      levels = rev(c("GLAC-1D · cold", "GLAC-1D · zonal", "ICE-6G · cold", "ICE-6G · zonal"))),
         region = factor(region, levels = rev(ice_order))) |>
  arrange(bar, desc(region)) |> group_by(bar) |>
  mutate(xmax = cumsum(share), xmin = xmax - share, xmid = (xmin + xmax) / 2) |> ungroup()
ggplot(nb) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = as.numeric(bar) - 0.38, ymax = as.numeric(bar) + 0.38, fill = region),
            colour = "white", linewidth = 0.4) +
  geom_text(data = nb |> filter(share >= 0.07), aes(xmid, as.numeric(bar), label = fmt(100 * share, 0)),
            size = 2.6, colour = "white", fontface = "bold") +
  scale_y_continuous(breaks = seq_along(levels(nb$bar)), labels = paste(levels(nb$bar), "mode"), expand = expansion(add = 0.3)) +
  scale_x_continuous(labels = function(x) paste0(100 * x, " %"), breaks = seq(0, 1, 0.25), expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = reg_col, breaks = ice_order, name = NULL) +
  labs(x = "Share of the transport-weighted anomaly at NISA", y = NULL) +
  theme_ms + theme(legend.position = "bottom", panel.grid.major.y = element_blank()) +
  guides(fill = guide_legend(nrow = 3))
```

## The Glas age-model ensemble

The rerun Bchron ensemble reproduces the published chronology
(@fig-agemodel). The ensemble median differs from the published ages by
`r fmt(sqrt(mean(dev^2)), 0)` yr root-mean-square and at most
`r fmt(max(abs(dev)), 0)` yr, at `r fmt(glas$depth_sample[which.max(abs(dev))], 1)` mm.
The median 95 % half-width is `r fmt(median(hw_ens), 0)` yr, against
`r fmt(median(hw_pub), 0)` yr for the published uncertainties, and it grows to
`r fmt(max(hw_ens), 0)` yr at the young end of the record, where the two
youngest dates are `r fmt(diff(dates$corr_age[1:2]), 0)` yr apart.

```{r}
#| label: fig-agemodel
#| fig-cap: "Glas age–depth model. Shading: 95 % (light) and 90 % (dark) range of the Bchron ensemble. Black line: published chronology. Points and bars: U–Th dates with their uncertainties."
#| fig-width: 6
#| fig-height: 3.6
agm <- tibble(depth = glas$depth_sample, pub = glas$interp_age,
              lo95 = ens_q["2.5%", ], hi95 = ens_q["97.5%", ], lo90 = ens_q["5%", ], hi90 = ens_q["95%", ])
ggplot(agm, aes(depth)) +
  geom_ribbon(aes(ymin = lo95 / 1000, ymax = hi95 / 1000), fill = "#cde2fb") +
  geom_ribbon(aes(ymin = lo90 / 1000, ymax = hi90 / 1000), fill = "#86b6ef") +
  geom_line(aes(y = pub / 1000), linewidth = 0.4) +
  geom_errorbar(data = dates, aes(x = depth_dating, ymin = (corr_age - corr_age_uncert_neg) / 1000,
                                  ymax = (corr_age + corr_age_uncert_pos) / 1000), width = 0.3, inherit.aes = FALSE) +
  geom_point(data = dates, aes(x = depth_dating, y = corr_age / 1000), size = 1.3, inherit.aes = FALSE) +
  labs(x = "Depth from top (mm)", y = "Age (ka BP)") + theme_ms
```

## Alignment between Glas δ¹⁸O and the discharge

```{r}
#| label: b1-prep
ksmooth_g <- function(t, x, h, tout = t) {
  W <- exp(-outer(tout, t, "-")^2 / (2 * h^2)); as.vector((W %*% x) / rowSums(W))
}
zs <- function(x) (x - mean(x)) / sd(x)
h_hp <- 750; h_k <- 25
kxcf <- function(tx, x, ty, y, lags, h) {     # tau > 0: y (discharge) leads x (Glas)
  D <- outer(tx, ty, "-"); XY <- outer(x, y)
  vapply(lags, function(tau) { W <- exp(-(D - tau)^2 / (2 * h^2)); sum(W * XY) / sum(W) }, numeric(1))
}
ar1_irreg <- function(t, tau_p, nsim) {
  n <- length(t); a <- exp(-abs(diff(t)) / tau_p)
  m <- matrix(rnorm(n * nsim), n, nsim)
  for (i in 2:n) m[i, ] <- a[i - 1] * m[i - 1, ] + sqrt(1 - a[i - 1]^2) * m[i, ]
  m
}
phase_rand <- function(y, nsim) {
  n <- length(y); f <- fft(y); half <- floor((n - 1) / 2)
  sapply(seq_len(nsim), function(s) {
    g <- f; g[2:(half + 1)] <- Mod(f[2:(half + 1)]) * exp(1i * runif(half, 0, 2 * pi))
    g[n:(n - half + 1)] <- Conj(g[2:(half + 1)]); Re(fft(g, inverse = TRUE)) / n
  })
}
tau_from <- function(t, x, st = 100) {
  b <- tibble(age = round(-t / st) * st, x = x) |> group_by(age) |> summarise(x = mean(x)) |> arrange(desc(age))
  phi <- cor(b$x[-1], b$x[-nrow(b)]); list(phi = phi, tau = -st / log(phi))
}

lags <- seq(-2000, 2000, 100)
tG <- -glas$interp_age
xG_hp <- zs(glas$d18O_measurement - ksmooth_g(tG, glas$d18O_measurement, h_hp))
xG_lv <- zs(glas$d18O_measurement)
wide <- long |> filter(recon == "GLAC-1D", age >= win[1] - 2500, age <= win[2] + 2500) |>
  select(age, region, anom) |> pivot_wider(names_from = region, values_from = anom) |> arrange(desc(age))
wide$Total <- rowSums(wide[, as.character(regions$region)])
tD <- -wide$age
inwin <- wide$age >= win[1] & wide$age <= win[2]
series <- c("Total", as.character(regions$region))
stdw <- function(y) (y - mean(y[inwin])) / sd(y[inwin])
Y_hp <- sapply(series, function(s) { y <- wide[[s]]; stdw(y - ksmooth_g(tD, y, h_hp)) })
Y_lv <- sapply(series, function(s) stdw(wide[[s]]))

p_hp <- tau_from(tG, xG_hp); p_lv <- tau_from(tG, xG_lv)
phi_d <- { y <- Y_hp[inwin, "Total"]; cor(y[-1], y[-length(y)]) }
n_eff <- sum(inwin) * (1 - p_hp$phi * phi_d) / (1 + p_hp$phi * phi_d)

obs_hp <- sapply(series, function(s) kxcf(tG, xG_hp, tD, Y_hp[, s], lags, h_k))
obs_lv <- sapply(series, function(s) kxcf(tG, xG_lv, tD, Y_lv[, s], 0, h_k))
```

```{r}
#| label: b1-null
NS <- 300
S_hp <- apply(ar1_irreg(tG, p_hp$tau, NS), 2, zs)
S_lv <- apply(ar1_irreg(tG, p_lv$tau, NS), 2, zs)
null_hp <- array(NA, c(length(lags), length(series), NS))
null_lv <- matrix(NA, length(series), NS)
for (k in seq_len(NS)) for (s in seq_along(series)) {
  null_hp[, s, k] <- kxcf(tG, S_hp[, k], tD, Y_hp[, series[s]], lags, h_k)
  null_lv[s, k]   <- kxcf(tG, S_lv[, k], tD, Y_lv[, series[s]], 0, h_k)
}
PR <- phase_rand(wide$Total - ksmooth_g(tD, wide$Total, h_hp), NS)
PR <- apply(PR, 2, stdw)
null_pr <- sapply(seq_len(NS), function(k) kxcf(tG, xG_hp, tD, PR[, k], lags, h_k))

env_hp <- apply(null_hp[, 1, ], 1, quantile, c(0.025, 0.975))
fw_all <- apply(abs(null_hp), 3, max)
fw_tot <- apply(abs(null_hp[, 1, ]), 2, max)
fw_pr  <- apply(abs(null_pr), 2, max)
p_fw_all <- mean(fw_all >= max(abs(obs_hp)))
p_fw_tot <- mean(fw_tot >= max(abs(obs_hp[, "Total"])))
p_fw_pr  <- mean(fw_pr  >= max(abs(obs_hp[, "Total"])))
p_hp0 <- sapply(seq_along(series), function(s) mean(abs(null_hp[lags == 0, s, ]) >= abs(obs_hp[lags == 0, s])))
p_lv0 <- sapply(seq_along(series), function(s) mean(abs(null_lv[s, ]) >= abs(obs_lv[s])))
obs_pk <- which.max(obs_hp[, "Total"])
```

```{r}
#| label: b1-ensemble
NE <- 200; idx <- round(seq(1, N, length.out = NE))
ens_hp <- array(NA, c(length(lags), length(series), NE)); ens_lv <- matrix(NA, length(series), NE)
for (j in seq_len(NE)) {
  tm <- -A[idx[j], ]
  xm <- zs(glas$d18O_measurement - ksmooth_g(tm, glas$d18O_measurement, h_hp))
  for (s in seq_along(series)) {
    ens_hp[, s, j] <- kxcf(tm, xm, tD, Y_hp[, series[s]], lags, h_k)
    ens_lv[s, j]   <- kxcf(tm, xG_lv, tD, Y_lv[, series[s]], 0, h_k)
  }
}
ens_pk_lag <- lags[apply(ens_hp[, 1, ], 2, which.max)]
ens_pk_r   <- apply(ens_hp[, 1, ], 2, max)
frac_fw    <- mean(apply(abs(ens_hp[, 1, ]), 2, max) > quantile(fw_tot, 0.95))
frac_fwall <- mean(apply(abs(ens_hp), 3, max) > quantile(fw_all, 0.95))
ens_hp0 <- ens_hp[lags == 0, , ]
q_lv <- apply(null_lv, 1, function(z) quantile(abs(z), 0.95))
frac_lv_sig <- rowMeans(abs(ens_lv) > q_lv)
```

**Correlation.** @tbl-cor lists the zero-lag correlations. On the full
(level) series, Glas correlates positively with the GLAC-1D total anomaly
(r = `r fmt(obs_lv[1])`), moderately positively with EIS_NorwegianSea
(r = `r fmt(obs_lv["EIS_NorwegianSea"])`) and strongly negatively with
EIS_BayOfBiscay (r = `r fmt(obs_lv["EIS_BayOfBiscay"])`). Bay of Biscay
discharge peaks at the Last Glacial Maximum, when Glas records its highest
δ¹⁸O. Against AR(1) surrogates with the high persistence of the level series,
`r ifelse(any(p_lv0 < 0.05), paste(paste(series[p_lv0 < 0.05], collapse = " and "), "reach"), "no series reaches")`
p < 0.05 (smallest p = `r fmtp(min(p_lv0))`, `r series[which.min(p_lv0)]`).
Across the age ensemble the level correlations barely change, because the age
uncertainty is small compared with millennial variations. They are stable,
but a shared millennial pattern of this length is also what red noise
produces often.

At centennial time scales the picture changes. The high-passed Glas series
correlates *negatively* with the high-passed total anomaly at zero lag
(r = `r fmt(obs_hp[lags == 0, "Total"])` on the published chronology,
`r rng(ens_hp0[1, ])` across the ensemble). No region reaches p < 0.05 at
zero lag (smallest p = `r fmtp(min(p_hp0))`, `r series[which.min(p_hp0)]`).
The high-passed series have lag-1 autocorrelations of
`r fmt(p_hp$phi)` (Glas, 100-yr bins) and `r fmt(phi_d)` (GLAC-1D total), which
leaves about `r fmt(n_eff, 0)` effectively independent pairs in
`r sum(inwin)` 100-yr steps.

```{r}
#| label: tbl-cor
#| tbl-cap: "Zero-lag kernel correlations between Glas δ¹⁸O and the GLAC-1D total and regional anomalies. Level: full series. High-pass: centennial residual after removing a 750-yr Gaussian smoother. r and p on the published chronology (p: two-sided, 300 AR(1) surrogates of Glas). Ensemble: 5–95 % range over 200 Bchron age models."
tibble(Series = ifelse(series == "Total", "Total (all regions)", series),
       `Level r` = fmt(obs_lv), `p` = fmtp(p_lv0),
       `Level r, ensemble` = apply(ens_lv, 1, rng),
       `High-pass r` = fmt(obs_hp[lags == 0, ]), `p ` = fmtp(p_hp0),
       `High-pass r, ensemble` = apply(ens_hp0, 1, rng)) |>
  knitr::kable()
```

**Lead and lag.** The high-passed cross-correlation with the GLAC-1D total
(@fig-ccf) peaks at a lag of `r lags[obs_pk]` yr (r = `r fmt(obs_hp[obs_pk, "Total"])`)
on the published chronology, which means Glas *leads* the discharge. Across
the ensemble the peak lag has a median of `r fmt(median(ens_pk_lag), 0)` yr
(5–95 %: `r fmt(quantile(ens_pk_lag, 0.05), 0)` to
`r fmt(quantile(ens_pk_lag, 0.95), 0)` yr) and a median peak r of
`r fmt(median(ens_pk_r))`. After accounting for the search over lags, the
peak is not significant (family-wise p = `r fmtp(p_fw_tot)` against AR(1)
surrogates of Glas and `r fmtp(p_fw_pr)` against phase-randomised discharge).
Over all ten series and all lags, family-wise p = `r fmtp(p_fw_all)`. The
family-wise 95 % threshold is exceeded by `r fmt(100 * frac_fw, 0)` % of
ensemble members for the total anomaly and by `r fmt(100 * frac_fwall, 0)` %
for any region.

```{r}
#| label: fig-ccf
#| fig-cap: "Kernel cross-correlation between high-passed Glas δ¹⁸O and the high-passed GLAC-1D total anomaly. Negative lags mean Glas leads. Line: published chronology. Blue band: 5–95 % range over 200 Bchron age models. Grey band: pointwise 95 % range of 300 AR(1) surrogates. Dashed lines: family-wise 95 % threshold over all lags."
#| fig-width: 6
#| fig-height: 3.2
ccf_df <- tibble(lag = lags, r = obs_hp[, "Total"], lo = env_hp[1, ], hi = env_hp[2, ],
                 elo = apply(ens_hp[, 1, ], 1, quantile, 0.05), ehi = apply(ens_hp[, 1, ], 1, quantile, 0.95))
fwq <- quantile(fw_tot, 0.95)
ggplot(ccf_df, aes(lag)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey88") +
  geom_ribbon(aes(ymin = elo, ymax = ehi), fill = "#86b6ef", alpha = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_hline(yintercept = c(-fwq, fwq), linetype = "dashed", linewidth = 0.3, colour = ink2) +
  geom_line(aes(y = r), colour = "#184f95", linewidth = 0.6) +
  labs(x = "Lag (yr); < 0: Glas leads", y = "Kernel correlation r") + theme_ms
```

```{r}
#| label: b2-events
ev_pub <- tribble(
  ~event,   ~mid_pub, ~lo,   ~hi,   ~sign, ~d1,   ~d2,
  "23.4kS", 23440, 23330, 24050,  1, 37.2, 39.2,
  "17.8kS", 17840, 17820, 18030, -1, 20.2, 20.95,
  "16.4kS", 16440, 16140, 16550, -1, 15.35, 15.5,
  "16.1kS", 16070, 15900, 16380, -1, 14.45, 14.75,
  "15.3kS", 15270, 15140, 15360, -1, 12.2, 12.8,
  "14.7kS", 14680, 14540, 14810,  1, 11.1, 11.45) |>
  mutate(depth = (d1 + d2) / 2, age_pub = approx(glas$depth_sample, glas$interp_age, depth)$y)
age_at_depth <- function(dep) sapply(dep, function(d) {
  i <- findInterval(d, glas$depth_sample, all.inside = TRUE)
  w <- (d - glas$depth_sample[i]) / (glas$depth_sample[i + 1] - glas$depth_sample[i])
  A[, i] * (1 - w) + A[, i + 1] * w })

tg <- tot |> filter(recon == "GLAC-1D", age >= win[1] - 600, age <= win[2] + 600) |>
  arrange(desc(age)) |> mutate(dA = tot - lag(tot), amid = age + 50)
sd_dA <- sd(tg$dA, na.rm = TRUE)
pick_runs <- function(idx, val, keep_min) {
  if (!length(idx)) return(integer(0))
  runs <- split(idx, cumsum(c(1, diff(idx) != 1)))
  vapply(runs, function(ix) ix[if (keep_min) which.min(val[ix]) else which.max(val[ix])], integer(1)) }
detect_pulses <- function(k) {
  thr <- k * sd_dA
  bind_rows(tibble(idx = pick_runs(which(tg$dA < -thr), tg$dA, TRUE), sign = -1),
            tibble(idx = pick_runs(which(tg$dA >  thr), tg$dA, FALSE), sign = 1)) |>
    mutate(age = tg$amid[idx], dA = tg$dA[idx]) |> arrange(desc(age)) }
nearest <- function(ages, signs, pl) sapply(seq_along(ages), function(i) {
  p <- pl$age[pl$sign == signs[i]]; if (length(p)) p[which.min(abs(p - ages[i]))] else NA })
match_count <- function(ages, signs, pl, tol) sum(abs(ages - nearest(ages, signs, pl)) <= tol, na.rm = TRUE)
null_rand <- function(signs, pl, tol, B = 5000)
  replicate(B, match_count(runif(length(signs), win[1], win[2]), signs, pl, tol))
ens_events <- function(EA, ev, pl, tol) {
  NM <- matrix(unlist(lapply(seq_len(nrow(EA)), function(m) nearest(EA[m, ], ev$sign, pl))), nrow = nrow(EA), byrow = TRUE)
  list(mc = rowSums(abs(EA - NM) <= tol, na.rm = TRUE), pm = colMeans(abs(EA - NM) <= tol, na.rm = TRUE), off = EA - NM)
}

pulses <- detect_pulses(1)
EA_pub <- age_at_depth(ev_pub$depth)
r_pub  <- ens_events(EA_pub, ev_pub, pulses, 300)
nr_pub <- null_rand(ev_pub$sign, pulses, 300)
mc_pub_med <- match_count(ev_pub$age_pub, ev_pub$sign, pulses, 300)
p_pub  <- mean(nr_pub >= median(r_pub$mc))
ev_pub <- ev_pub |> mutate(pulse = nearest(age_pub, sign, pulses), offset = age_pub - pulse,
                           pm = r_pub$pm, ens_lo = apply(EA_pub, 2, quantile, 0.05),
                           ens_hi = apply(EA_pub, 2, quantile, 0.95),
                           off_lo = apply(r_pub$off, 2, quantile, 0.05), off_hi = apply(r_pub$off, 2, quantile, 0.95))
ev_hw <- (ev_pub$ens_hi - ev_pub$ens_lo) / 2

sens <- expand_grid(k = c(1, 1.5, 2), tol = c(200, 300, 500)) |> rowwise() |>
  mutate(pl = list(detect_pulses(k)), n_pulses = nrow(pl),
         mc = list(ens_events(EA_pub, ev_pub, pl, tol)$mc),
         mc_med = median(mc), mc_lo = quantile(mc, 0.05), mc_hi = quantile(mc, 0.95),
         nl = list(null_rand(ev_pub$sign, pl, tol, 2000)),
         null_mean = mean(nl), p = mean(nl >= mc_med), p_frac = mean(sapply(mc, function(m) mean(nl >= m)) < 0.05)) |>
  ungroup() |> select(-pl, -mc, -nl)

deltas <- seq(-1000, 1000, 100)
scan <- map_dfr(deltas, function(dl) {
  pl <- pulses |> mutate(age = age + dl); mc <- ens_events(EA_pub, ev_pub, pl, 300)$mc
  tibble(delta = dl, pub = match_count(ev_pub$age_pub, ev_pub$sign, pl, 300),
         med = median(mc), lo = quantile(mc, 0.05), hi = quantile(mc, 0.95)) })

# Objective breakpoints on the 100-yr binned series, calibrated against AR(1) surrogates
gb <- glas |> mutate(age = round(interp_age / 100) * 100) |> group_by(age) |>
  summarise(d = mean(d18O_measurement), depth = mean(depth_sample)) |> arrange(desc(age))
rss_gain <- function(y, kmax = 10, h = 3) {
  bpo <- breakpoints(y ~ 1, h = h, breaks = kmax); r <- summary(bpo)$RSS["RSS", ]
  list(bp = bpo, gain = 1 - r[-1] / r[-length(r)]) }
obs_bp <- rss_gain(gb$d)
phi_lvb <- cor(gb$d[-1], gb$d[-nrow(gb)])
sur_gain <- replicate(200, rss_gain(as.vector(arima.sim(list(ar = phi_lvb), n = nrow(gb))))$gain)
q_gain <- apply(sur_gain, 1, quantile, 0.95)
n_bp <- sum(cumprod(obs_bp$gain > q_gain))
bpi <- breakpoints(obs_bp$bp, breaks = n_bp)$breakpoints
seg <- c(0, bpi, nrow(gb))
segmean <- sapply(seq_len(length(seg) - 1), function(i) mean(gb$d[(seg[i] + 1):seg[i + 1]]))
ev_obj <- tibble(age_pub = (gb$age[bpi] + gb$age[bpi + 1]) / 2, depth = (gb$depth[bpi] + gb$depth[bpi + 1]) / 2,
                 jump = segmean[-1] - segmean[-length(segmean)], sign = sign(jump))
EA_obj <- age_at_depth(ev_obj$depth)
r_obj <- ens_events(EA_obj, ev_obj, pulses, 300)
ev_obj$pulse <- nearest(ev_obj$age_pub, ev_obj$sign, pulses)
```

**Event matching.** Within the Glas interval, the GLAC-1D total anomaly has
`r sum(pulses$sign == -1)` freshening pulses and `r sum(pulses$sign == 1)`
recoveries above the 1-SD threshold of `r fmt(sd_dA, 1)` ‰ per 100 yr
(@fig-series). On the published chronology, `r mc_pub_med` of the six
published Glas events have a same-sign pulse within ±300 yr (@tbl-events).
Across the age ensemble the count ranges from `r min(r_pub$mc)` to
`r max(r_pub$mc)` (median `r median(r_pub$mc)`). Random placement of six
events gives `r fmt(mean(nr_pub), 1)` matches on average, so the median count
has p = `r fmtp(p_pub)`. The Glas age uncertainty at the events
(`r fmt(min(ev_hw), 0)`–`r fmt(max(ev_hw), 0)` yr, 5–95 % half-width) is
comparable to the matching tolerance, so the match status of individual
events is itself uncertain:

- **Robust matches.** `r paste(ev_pub$event[ev_pub$pm >= 0.8], collapse = ", ")`
  are matched in at least 80 % of age models
  (`r paste0(fmt(100 * ev_pub$pm[ev_pub$pm >= 0.8], 0), " %", collapse = ", ")`).
  These are the Heinrich Stadial 1 freshening steps and the
  Last Glacial Maximum re-salinisation.
- **Uncertain matches.** `r paste(ev_pub$event[ev_pub$pm < 0.8], collapse = " and ")`
  are matched in only `r paste0(fmt(100 * ev_pub$pm[ev_pub$pm < 0.8], 0), " %", collapse = " and ")`
  of age models. The 17.8kS freshening onset lies
  `r fmt(-ev_pub$offset[ev_pub$event == "17.8kS"], 0)` yr after the GLAC-1D
  freshening pulse at `r ka(ev_pub$pulse[ev_pub$event == "17.8kS"], 2)` ka.
  The 14.7kS re-salinisation lies `r fmt(ev_pub$offset[ev_pub$event == "14.7kS"], 0)` yr
  before the nearest recovery. The nearest GLAC-1D *freshening* pulse lies
  only `r fmt(abs(ev_pub$age_pub[6] - nearest(ev_pub$age_pub[6], -1, pulses)), 0)` yr
  away, at `r ka(nearest(ev_pub$age_pub[6], -1, pulses), 2)` ka, during Meltwater
  Pulse 1A. The event therefore aligns better with discharge of the opposite
  sign.

@fig-matches shows the six events on the Glas record, together with the
nearest same-sign GLAC-1D pulse and how often the match holds across the age
ensemble.

```{r}
#| label: fig-matches
#| fig-cap: "Which abrupt Glas shifts match a GLAC-1D discharge pulse. Grey points: Glas δ¹⁸O samples; black line: 100-yr means. Bands: the six published events, spanning their 5–95 % ensemble age range (blue freshening, orange re-salinisation); solid bands are robust matches (same-sign pulse within ±300 yr in ≥ 80 % of age models), pale dashed bands are uncertain. Triangles at the bottom: GLAC-1D pulses above 1 SD (down: freshening, up: recovery). The line from each event to its nearest same-sign pulse is labelled with the offset on the published chronology and the match probability P."
#| fig-width: 7
#| fig-height: 4.2
yr <- range(glas$d18O_measurement)
y_pulse <- yr[1] - 0.75; y_link0 <- yr[1] - 0.25; y_lab0 <- yr[2] + 0.2
mdf <- ev_pub |> mutate(robust = pm >= 0.8, col = ifelse(sign < 0, col_fresh, col_resal),
                        lab = sprintf("%s\nP = %s", event, fmt(pm, 2)),
                        olab = sprintf("%+d yr", round(offset)),
                        lvl = (seq_len(n()) - 1) %% 2,            # alternate two label levels
                        y_lab = y_lab0 + 0.42 * lvl, y_link = y_link0 - 0.25 * lvl)
pl_win <- pulses |> filter(age >= win[1] - 300, age <= win[2] + 300)
ggplot() +
  geom_rect(data = mdf, aes(xmin = ens_lo / 1000, xmax = ens_hi / 1000, ymin = -Inf, ymax = Inf, fill = col,
                            alpha = robust, linetype = robust), colour = NA) +
  geom_rect(data = mdf |> filter(!robust), aes(xmin = ens_lo / 1000, xmax = ens_hi / 1000, ymin = y_link0 + 0.1, ymax = yr[2] + 0.1,
                                               colour = col), fill = NA, linetype = "22", linewidth = 0.35) +
  geom_point(data = glas, aes(interp_age / 1000, d18O_measurement), size = 0.35, colour = "grey65") +
  geom_line(data = gb, aes(age / 1000, d), linewidth = 0.45) +
  geom_point(data = pl_win, aes(age / 1000, y_pulse, shape = factor(sign), colour = ifelse(sign < 0, col_fresh, col_resal)),
             size = 2, fill = NA) +
  geom_segment(data = mdf, aes(x = age_pub / 1000, xend = pulse / 1000, y = y_link, yend = y_link, colour = col),
               linewidth = 0.6, arrow = arrow(length = unit(3, "pt"), type = "closed")) +
  geom_text(data = mdf, aes((age_pub + pulse) / 2000, y_link + 0.13, label = olab), size = 2.3, colour = "grey25") +
  geom_text(data = mdf, aes(age_pub / 1000, y_lab, label = lab, fontface = ifelse(robust, "bold", "plain")),
            size = 2.4, lineheight = 0.9, vjust = 0) +
  scale_fill_identity() + scale_colour_identity() +
  scale_alpha_manual(values = c(`TRUE` = 0.35, `FALSE` = 0.1), guide = "none") +
  scale_linetype_manual(values = c(`TRUE` = "solid", `FALSE` = "22"), guide = "none") +
  scale_shape_manual(values = c("-1" = 25, "1" = 24), guide = "none") +
  scale_x_reverse(breaks = seq(12, 24, 1), limits = c(win[2] / 1000 + 0.35, win[1] / 1000 - 0.3)) +
  scale_y_continuous(limits = c(y_pulse - 0.15, y_lab0 + 0.42 + 0.5), breaks = seq(-5, -1, 0.5)) +
  labs(x = "Age (ka BP)", y = expression("Glas" ~ delta^18*O ~ "(‰)")) + theme_ms
```

@tbl-sens shows that significance depends on the tolerance. At ±200–300 yr
no combination reaches p < 0.05. At ±500 yr the median count is significant
for every threshold (p = `r paste(fmtp(sens$p[sens$tol == 500]), collapse = ", ")`).
The offset scan (@fig-scan) shows that shifting the discharge by a few hundred
years changes the median count between `r min(scan$med)` and `r max(scan$med)`,
with no single offset clearly preferred. The alignment is therefore
detectable at the ~500-yr scale, which is also the scale of the combined
chronological uncertainty, but not at the centennial scale.

**Objective events.** Only `r n_bp` mean-shift breakpoint in Glas δ¹⁸O
exceeds what AR(1) noise produces: a
`r ifelse(ev_obj$sign[1] < 0, "freshening", "re-salinisation")` step of
`r fmt(ev_obj$jump[1])` ‰ at `r ka(ev_obj$age_pub[1], 2)` ka, the published
17.8kS onset. It lies `r fmt(abs(ev_obj$age_pub[1] - ev_obj$pulse[1]), 0)` yr
from the nearest GLAC-1D freshening pulse and is matched within ±300 yr in
`r fmt(100 * r_obj$pm[1], 0)` % of age models. The other published events are
real features of Glas, but none improves the segmentation more than red noise
with the same persistence would. The surrogate test is
conservative, because the persistence estimate includes the steps themselves.

```{r}
#| label: tbl-events
#| tbl-cap: "Published abrupt δ¹⁸O events in Glas (Endres et al., 2026a, Table 1) and the nearest same-sign GLAC-1D pulse. Sign: −1 freshening, +1 re-salinisation. Age and offset on the published chronology, with the 5–95 % range over the Bchron ensemble in brackets. Offset: Glas age minus pulse age (positive: Glas earlier). P(match): fraction of age models with |offset| ≤ 300 yr. Where the offset range spans both signs, the nearest pulse changes between age models."
ev_pub |> transmute(Event = event, Sign = sign, `Depth (mm)` = fmt(depth, 2),
                    `Age (ka)` = paste0(ka(age_pub, 2), " (", ka(ens_lo, 2), "–", ka(ens_hi, 2), ")"),
                    `Pulse (ka)` = ka(pulse, 2),
                    `Offset (yr)` = paste0(fmt(offset, 0), " (", fmt(off_lo, 0), " to ", fmt(off_hi, 0), ")"),
                    `P(match)` = fmt(pm, 2)) |>
  knitr::kable()
```

```{r}
#| label: tbl-sens
#| tbl-cap: "Sensitivity of event matching (six published events) to the pulse threshold (k SD of 100-yr changes in the GLAC-1D total anomaly) and the tolerance. Matched: median and 5–95 % range over the age ensemble. Null mean and p from 2,000 random placements. Sig. models: fraction of age models with p < 0.05."
sens |> transmute(`k (SD)` = k, `Tolerance (yr)` = tol, Pulses = n_pulses,
                  `Matched (of 6)` = paste0(mc_med, " (", mc_lo, "–", mc_hi, ")"),
                  `Null mean` = fmt(null_mean, 1), p = fmtp(p), `Sig. models (%)` = fmt(100 * p_frac, 0)) |>
  knitr::kable()
```

```{r}
#| label: fig-series
#| fig-cap: "Glas δ¹⁸O (top; grey points on the published chronology, black line 100-yr means) and the total source anomaly of GLAC-1D (middle) and ICE-6G (bottom). Shaded bars: published events, spanning the 5–95 % ensemble age range (blue freshening, orange re-salinisation). Triangles: GLAC-1D pulses above 1 SD (down: freshening, up: recovery)."
#| fig-width: 7
#| fig-height: 6.2
xl <- c(win[2] / 1000 + 0.4, win[1] / 1000 - 0.4)
evr <- geom_rect(data = ev_pub, aes(xmin = ens_lo / 1000, xmax = ens_hi / 1000, ymin = -Inf, ymax = Inf,
                                    fill = factor(sign)), alpha = 0.35, inherit.aes = FALSE)
fl <- scale_fill_manual(values = c("-1" = col_fresh, "1" = col_resal), guide = "none")
xs <- scale_x_reverse(limits = xl, breaks = seq(12, 24, 2))
p1 <- ggplot() + evr + fl +
  geom_point(data = glas, aes(interp_age / 1000, d18O_measurement), size = 0.4, colour = "grey60") +
  geom_line(data = gb, aes(age / 1000, d), linewidth = 0.4) + xs +
  labs(x = NULL, y = expression("Glas" ~ delta^18*O ~ "(‰)")) + theme_ms
tg_w <- tot |> filter(recon == "GLAC-1D", age >= win[1] - 400, age <= win[2] + 400)
p2 <- ggplot() + evr + fl +
  geom_line(data = tg_w, aes(age / 1000, tot), linewidth = 0.4) +
  geom_point(data = pulses |> filter(age >= win[1] - 400, age <= win[2] + 400),
             aes(age / 1000, ifelse(sign < 0, min(tg_w$tot) - 1.5, max(tg_w$tot) + 1.5), shape = factor(sign)),
             size = 1.8, fill = "black") +
  scale_shape_manual(values = c("-1" = 25, "1" = 24), guide = "none") + xs +
  labs(x = NULL, y = "GLAC-1D anomaly (‰)") + theme_ms
p3 <- ggplot() + evr + fl +
  geom_line(data = tot |> filter(recon == "ICE-6G", age >= win[1] - 500, age <= win[2] + 500),
            aes(age / 1000, tot), linewidth = 0.4) + xs +
  labs(x = "Age (ka BP)", y = "ICE-6G anomaly (‰)") + theme_ms
p1 / p2 / p3
```

```{r}
#| label: fig-scan
#| fig-cap: "Number of the six published Glas events matched by a same-sign GLAC-1D pulse within ±300 yr when the discharge series is shifted in time (positive: discharge moved older). Points: published chronology. Line and band: median and 5–95 % range over the age ensemble. Dashed line: mean of the random-placement null."
#| fig-width: 5.5
#| fig-height: 3
ggplot(scan, aes(delta)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#cde2fb") +
  geom_line(aes(y = med), colour = "#184f95", linewidth = 0.6) +
  geom_point(aes(y = pub), size = 1.4) +
  geom_hline(yintercept = mean(nr_pub), linetype = "dashed", colour = ink2, linewidth = 0.3) +
  scale_y_continuous(limits = c(0, 6), breaks = 0:6) +
  labs(x = "Shift of discharge series (yr)", y = "Events matched (of 6)") + theme_ms
```

```{r}
#| label: b3-sea
rel_bins <- seq(-500, 1000, 100); rb <- rel_bins[-length(rel_bins)]
sea_one <- function(ages_m, pl_ages) {
  comp <- sapply(pl_ages, function(ap) {
    b <- findInterval(ap - ages_m, rel_bins); ok <- b >= 1 & b <= length(rb)
    v <- tapply(glas$d18O_measurement[ok], factor(b[ok], levels = seq_along(rb)), mean)
    v - mean(v[rb < 0], na.rm = TRUE) })
  rowMeans(comp, na.rm = TRUE)
}
sea_stat <- function(comp) mean(comp[rb >= 100 & rb <= 500])
pf <- pulses$age[pulses$sign == -1 & pulses$age > win[1] + 500 & pulses$age < win[2] - 1000]
pr <- pulses$age[pulses$sign ==  1 & pulses$age > win[1] + 500 & pulses$age < win[2] - 1000]
sea_f <- sapply(seq_len(N), function(m) sea_one(A[m, ], pf))
sea_r <- sapply(seq_len(N), function(m) sea_one(A[m, ], pr))
null_comp <- function(np, B = 2000) replicate(B, sea_one(glas$interp_age, runif(np, win[1] + 500, win[2] - 1000)))
nc_f <- null_comp(length(pf)); nc_r <- null_comp(length(pr))
sf <- apply(sea_f, 2, sea_stat); sr <- apply(sea_r, 2, sea_stat)
nf <- apply(nc_f, 2, sea_stat); nr <- apply(nc_r, 2, sea_stat)
p_sf <- mean(nf <= median(sf)); p_sr <- mean(nr >= median(sr))
```

**Superposed epoch analysis.** Composited around the
`r length(pf)` GLAC-1D freshening pulses, Glas δ¹⁸O changes by
`r fmt(median(sf), 3)` ‰ on average 100–600 yr after the pulse (ensemble
5–95 %: `r rng(sf, 3)` ‰), indistinguishable from random pulse times
(one-sided p = `r fmtp(p_sf)`; @fig-sea). Around the `r length(pr)`
recoveries Glas changes by `r fmt(median(sr), 3)` ‰ (`r rng(sr, 3)` ‰), the
opposite of the expected re-salinisation (p = `r fmtp(p_sr)` for a positive
response). On average, the GLAC-1D pulses leave no detectable imprint in Glas
at centennial scale. The robust event matches above are therefore specific
coincidences, not a systematic response.

```{r}
#| label: fig-sea
#| fig-cap: "Superposed epoch composites of Glas δ¹⁸O around GLAC-1D freshening pulses (left) and recoveries (right), relative to the 500-yr pre-pulse mean. Line and blue band: median and 5–95 % range over the age ensemble. Grey band: 5–95 % range for random pulse times."
#| fig-width: 6.5
#| fig-height: 2.8
sea_df <- bind_rows(
  tibble(type = "Freshening pulses", t = rb + 50, med = apply(sea_f, 1, median),
         lo = apply(sea_f, 1, quantile, 0.05), hi = apply(sea_f, 1, quantile, 0.95),
         nlo = apply(nc_f, 1, quantile, 0.05, na.rm = TRUE), nhi = apply(nc_f, 1, quantile, 0.95, na.rm = TRUE)),
  tibble(type = "Recoveries", t = rb + 50, med = apply(sea_r, 1, median),
         lo = apply(sea_r, 1, quantile, 0.05), hi = apply(sea_r, 1, quantile, 0.95),
         nlo = apply(nc_r, 1, quantile, 0.05, na.rm = TRUE), nhi = apply(nc_r, 1, quantile, 0.95, na.rm = TRUE)))
ggplot(sea_df, aes(t)) +
  geom_ribbon(aes(ymin = nlo, ymax = nhi), fill = "grey88") +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#86b6ef", alpha = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.3) + geom_vline(xintercept = 0, linewidth = 0.3, linetype = "dashed") +
  geom_line(aes(y = med), colour = "#184f95", linewidth = 0.6) +
  facet_wrap(~type) + labs(x = "Time after pulse (yr)", y = expression("Glas" ~ Delta*delta^18*O ~ "(‰)")) + theme_ms
```

```{r}
#| label: b4-detect
Dg <- approx(tD, Y_hp[, "Total"], xout = tG)$y
snr <- c(0.05, 0.1, 0.2, 0.3, 0.5)
det <- sapply(snr, function(s) {
  e <- ar1_irreg(tG, p_hp$tau, 100)
  apply(e, 2, function(z) kxcf(tG, zs(sqrt(s) * Dg + sqrt(1 - s) * z), tD, Y_hp[, "Total"], 0, h_k)) })
r_obs0 <- obs_hp[lags == 0, "Total"]
frac_below <- colMeans(det <= quantile(ens_hp0[1, ], 0.95))
```

**Detectability.** If the high-passed GLAC-1D total explained a fraction of
the centennial Glas variance, the zero-lag correlation would be positive
(@tbl-detect). A signal fraction of 10 % already gives a median r of
`r fmt(median(det[, 2]))`. The observed ensemble range
(`r rng(ens_hp0[1, ])`) lies below the 5th percentile of the synthetic
correlations for every signal fraction of `r fmt(100 * snr[min(which(apply(det, 2, quantile, 0.05) > quantile(ens_hp0[1, ], 0.95)))], 0)` % and more. At zero lag, the
GLAC-1D forcing is therefore not a detectable component of centennial Glas
variability, even allowing for the age uncertainty.

```{r}
#| label: tbl-detect
#| tbl-cap: "Zero-lag kernel correlation expected between synthetic Glas series (signal fraction of the variance from the high-passed GLAC-1D total, remainder AR(1) noise with Glas persistence) and the GLAC-1D total, from 100 realisations each. Observed: 5–95 % range over the age ensemble."
tibble(`Signal fraction (%)` = c(fmt(100 * snr, 0), "Observed"),
       `Median r` = c(fmt(apply(det, 2, median)), fmt(median(ens_hp0[1, ]))),
       `5–95 % range` = c(apply(det, 2, rng), rng(ens_hp0[1, ]))) |>
  knitr::kable()
```

```{r}
#| label: b5-ice6g
i6 <- tot |> filter(recon == "ICE-6G", age >= win[1], age <= win[2]) |> arrange(desc(age))
g500b <- glas |> mutate(age = round(interp_age / 500) * 500) |> group_by(age) |> summarise(d = mean(d18O_measurement))
j6 <- inner_join(i6, g500b, by = "age")
r6_lv <- cor(j6$tot, j6$d); r6_df <- cor(diff(j6$tot), diff(j6$d))
```

**ICE-6G.** With `r nrow(j6)` 500-yr steps in the Glas interval, Glas
correlates with the ICE-6G total at r = `r fmt(r6_lv)` on levels, but at
r = `r fmt(r6_df)` on 500-yr changes. The level correlation reflects the
shared deglacial trend. ICE-6G ramps up smoothly towards Meltwater Pulse 1A
and has no centennial structure to align with the Glas events (@fig-series).

# Answer

**Most important source region.** It depends on the ice-sheet
reconstruction. Within the Glas interval, GLAC-1D has three nearly equal
contributors, `r top3$region[1]`, `r top3$region[2]` and `r top3$region[3]`
(`r fmt(100 * top3$share[3], 0)`–`r fmt(100 * top3$share[1], 0)` % each of the
summed source anomaly), with European sectors supplying
`r gs("GLAC-1D", "European")` %. ICE-6G gives
`r imp(s_win, "ICE-6G")$region` (`r fmt(100 * imp(s_win, "ICE-6G")$share, 0)` %),
with Laurentide sectors supplying `r gs("ICE-6G", "Laurentide")` %. Over the
full series, both reconstructions put `r imp(s_full, "GLAC-1D")$region` first.
Weighted by transport to NISA, the answer shifts toward regions that reach the
cave efficiently. In GLAC-1D, `r nt("GLAC-1D", "cold")$region` becomes the
most important region at NISA (`r fmt(100 * min(nisa_top$share[nisa_top$recon == "GLAC-1D"]), 0)`–`r fmt(100 * max(nisa_top$share[nisa_top$recon == "GLAC-1D"]), 0)` %
of the site anomaly in the two AMOC modes). In ICE-6G it remains
`r nt("ICE-6G", "cold")$region`
(`r fmt(100 * min(nisa_top$share[nisa_top$recon == "ICE-6G"]), 0)`–`r fmt(100 * max(nisa_top$share[nisa_top$recon == "ICE-6G"]), 0)` %).

**Most variable source region.** In GLAC-1D it is
`r var_(s_win, "GLAC-1D")$region`, which has the largest SD and
`r fmt(100 * vsh(s_win, "GLAC-1D")$vshare, 0)` % of the variance of the total
anomaly. In ICE-6G, `r var_(s_win, "ICE-6G")$region` and
`r vsh(s_win, "ICE-6G")$region` share this role. In both reconstructions the
variability comes from a few short pulses. At NISA, after transport
weighting, the most variable contribution is
`r nt("GLAC-1D", "cold", nisa_var)$region` in GLAC-1D and
`r nt("ICE-6G", "cold", nisa_var)$region` in ICE-6G.

**Alignment with Glas.** Glas and the discharge share millennial structure
and several individual events. There is no systematic centennial response.

- *Levels.* Glas follows the millennial evolution of EIS_NorwegianSea
  (r = `r fmt(obs_lv["EIS_NorwegianSea"])`) and anti-follows EIS_BayOfBiscay
  (r = `r fmt(obs_lv["EIS_BayOfBiscay"])`). Both are robust to age uncertainty,
  but neither is significant against red noise
  (p = `r fmtp(p_lv0[series == "EIS_NorwegianSea"])` and `r fmtp(p_lv0[series == "EIS_BayOfBiscay"])`).
- *Centennial correlation.* High-passed Glas and GLAC-1D are not positively
  correlated at zero lag (ensemble `r rng(ens_hp0[1, ])`). The best lead–lag
  has Glas leading by `r fmt(-lags[obs_pk], 0)` yr on the published chronology
  (`r fmt(-median(ens_pk_lag), 0)` yr ensemble median), but it is not significant once the search over lags is accounted for
  (family-wise p = `r fmtp(p_fw_tot)`).
- *Events.* `r median(r_pub$mc)` of the six published abrupt events have a
  same-sign GLAC-1D pulse within ±300 yr (ensemble range
  `r min(r_pub$mc)`–`r max(r_pub$mc)`; p = `r fmtp(p_pub)`). The count becomes
  significant at ±500 yr (p ≤ `r fmtp(max(sens$p[sens$tol == 500]))`). The
  three Heinrich Stadial 1 freshening steps and the 23.4 ka re-salinisation
  match in ≥ 80 % of age models. The 17.8 ka onset, the only step that
  exceeds red noise, follows a GLAC-1D freshening pulse by
  ~`r fmt(round(-ev_pub$offset[ev_pub$event == "17.8kS"], -1), 0)` yr. The 14.7 ka
  re-salinisation coincides with Meltwater Pulse 1A, the opposite sign.

**How to quantify it.** The appropriate quantification is event-based and
carried through an age-model ensemble: per-event match probabilities and
offsets, a match count tested against random placement, and a
superposed-epoch composite that tests for a systematic response. Correlation
of the full series mainly measures the shared trend. Correlations of
high-passed series must be tested with a family-wise null over lags and
regions and with surrogates for both series. With a ±150–200 yr age
uncertainty in Glas and an unquantified uncertainty in the ice-sheet
chronologies, centennial alignment cannot be demonstrated from these data;
alignment at the ~500-yr scale can.

# Caveats

- **Equilibrium transport, one AMOC mode at a time.** The alignment tests
  use the source-region forcing. The site anomalies use equilibrium
  transport weights (Endres et al., 2026b), which ignore transit times and
  hold the AMOC mode fixed. The forward-model draft (in prep.) switches modes
  through the deglaciation and reproduces the 14.7 ka case as AMOC recovery
  cancelling the Meltwater Pulse 1A anomaly at NISA, which neither approach
  here can capture. The most direct test would compare Glas with the
  convolved NISA series, which is not in the repository.
- **Discharge chronology.** The ice-sheet reconstructions have their own age
  uncertainty, which is only bracketed here by the offset scan.
- **Glas δ¹⁸O is not only seawater.** It also records temperature, moisture
  source and the global-ocean deglacial trend, which lowers the expected
  correlation.
- **Ensemble sizes.** `r fmt(N, 0)` Bchron age models were used for event statistics
  and 200 for cross-correlations. Surrogate counts of 300 limit p values to
  a resolution of about 0.003.

# References

Endres, L., Pérez-Mejías, C., Ivanovic, R., Gregoire, L., Hughes, A. L. C.,
Cheng, H., and Stoll, H. (2026a). Interplay of North Atlantic freshening and
deep convection during the last deglaciation constrained by Iberian
speleothems. *Climate of the Past*, 22, 797–824.
https://doi.org/10.5194/cp-22-797-2026

Endres, L., Ivanovic, R., Rome, Y., Tindall, J., Thurnherr, I., Sprenger, M.,
and Stoll, H. (2026b). Tracing meltwater from northern ice sheets to
palaeoclimate archives during the early last deglaciation: a conservative
tracer approach. *ESS Open Archive* (preprint).
https://doi.org/10.22541/essoar.15007256/v1

Endres, L., Ivanovic, R., Rome, Y., and Stoll, H. (in prep.). A dye tracer
forward-model for meltwater-related surface ocean anomalies. Draft
(`Impulse_Response_Revisions_v1.pdf`, this repository).

Franke, J. G., and Donner, R. V. (2019). Correlating paleoclimate time series:
sources of uncertainty and potential pitfalls. *Quaternary Science Reviews*.
Preprint: https://arxiv.org/abs/1903.11865

Haslett, J., and Parnell, A. (2008). A simple monotone process with
application to radiocarbon-dated depth chronologies. *Journal of the Royal
Statistical Society C*, 57, 399–418.

Rehfeld, K., and Kurths, J. (2014). Similarity estimators for irregular and
age-uncertain time series. *Climate of the Past*, 10, 107–122.
https://doi.org/10.5194/cp-10-107-2014

Rehfeld, K., Marwan, N., Heitzig, J., and Kurths, J. (2011). Comparison of
correlation analysis techniques for irregularly sampled time series.
*Nonlinear Processes in Geophysics*, 18, 389–404.
https://doi.org/10.5194/npg-18-389-2011

Schulz, M., and Mudelsee, M. (2002). REDFIT: estimating red-noise spectra
directly from unevenly spaced paleoclimatic time series. *Computers &
Geosciences*, 28, 421–426.

Theiler, J., Eubank, S., Longtin, A., Galdrikian, B., and Farmer, J. D.
(1992). Testing for nonlinearity in time series: the method of surrogate
data. *Physica D*, 58, 77–94.

Zeileis, A., Leisch, F., Hornik, K., and Kleiber, C. (2002). strucchange: an R
package for testing for structural change in linear regression models.
*Journal of Statistical Software*, 7(2), 1–38.
