---
title: "Where the meltwater arrives: source-region contributions at nine proxy sites (forward model v2)"
author: "Laura Endres"
date: today
format:
  docx:
    toc: false
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("R/site_metrics.R")
library(purrr); library(jsonlite)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 0) fmt(100 * x, d)

win  <- c(21500, 10000)
d    <- read_site_anomaly()
g    <- to_grid(d, step = 500, window = win)
m    <- site_metrics(g)
info <- fromJSON("outputs/convolution/run_info.json")
chk  <- read_csv("outputs/convolution/scaling_check.csv", show_col_types = FALSE)
val  <- read_csv("outputs/convolution/validation_constant_forcing.csv", show_col_types = FALSE)

sites <- c("La Vallina (NISA)", "Cave Without a Name", "NGRIP", "MD95-2010", "PS2644-5",
           "MD03-2664", "ODP980", "MD95-2042", "MD02-2552")
short <- c(EIS_MedSea = "Med", EIS_BayOfBiscay = "Bri", EIS_NorwegianSea = "Fen", EIS_Arctic = "EurArc",
           LAU_Arctic = "AmeArc", GIS_GreenlandSea = "GIS", LAU_LabradorSea = "NLau",
           LAU_StLawrence = "SLau", LAU_GulfOfMexico = "GoM")
theme_ms <- theme_bw(base_size = 9) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_line(linewidth = 0.2, colour = "grey90"),
        strip.background = element_blank(), strip.text = element_text(face = "bold", hjust = 0))

ss   <- site_summary(m)
mix  <- ss |> filter(pathway == "mixed")
top  <- function(f, s, what = "important") mix[[what]][mix$forcing == f & mix$site == s]
topv <- function(f, s, what = "share") mix[[what]][mix$forcing == f & mix$site == s]
shr  <- function(s) { v <- sort(mix$share[mix$site == s]); paste0(pct(v[1]), "–", pct(v[2])) }
agree <- mix |> select(forcing, site, important) |> pivot_wider(names_from = forcing, values_from = important) |>
  mutate(same = `GLAC-1D` == `ICE-6G`)
n_agree <- sum(agree$same)
n_fen_abrupt <- sum(mix$abrupt == "EIS_NorwegianSea")
robust <- ss |> group_by(forcing, site) |> summarise(same = n_distinct(important) == 1, .groups = "drop")
n_robust <- sum(robust$same)

# Worked example for the metric explanations: NISA, GLAC-1D, mixed pathway.
ex <- m |> filter(site == "La Vallina (NISA)", forcing == "GLAC-1D", pathway == "mixed")
ex_top <- ex |> slice_max(share, n = 1)
ex_dom <- ex |> slice_max(dom, n = 1)
ex_sd  <- ex |> slice_max(sd, n = 1)
ex_vs  <- ex |> slice_max(vshare, n = 1)
ex_ab  <- ex |> slice_max(sd_diff, n = 1)

# Kernel checks (cold and zonal can be compared with the paper's equilibrium weights).
w_site <- chk |> filter(!is.na(equilibrium_weight)) |> group_by(site, mode) |>
  summarise(r = sum(response_500yr) / sum(equilibrium_weight), .groups = "drop")
```

# Summary

The meltwater anomaly at each of the nine proxy sites of Endres et al.
(2026b) was computed with the forward model: the regional meltwater forcing
of GLAC-1D or ICE-6G, convolved with the HadCM3 response to a dye pulse from
each of the nine input regions. This replaces the equilibrium approximation
of the previous report with the real pulse responses, so meltwater now
arrives late and spread out, as it does in the ocean model.

- The two ice-sheet reconstructions agree on the most important source region
  at `r n_agree` of the 9 sites, in the mixed AMOC pathway, 21.5–10 ka.
- At La Vallina (NISA), GLAC-1D makes `r top("GLAC-1D", "La Vallina (NISA)")`
  the most important source (`r pct(topv("GLAC-1D", "La Vallina (NISA)"))` % of
  the anomaly). ICE-6G makes it `r top("ICE-6G", "La Vallina (NISA)")`
  (`r pct(topv("ICE-6G", "La Vallina (NISA)"))` %).
- The two sites near the Gulf of Mexico are dominated by LAU_GulfOfMexico in
  both reconstructions: `r shr("MD02-2552")` % of the anomaly at MD02-2552 and
  `r shr("Cave Without a Name")` % at Cave Without a Name (Texas).
- EIS_NorwegianSea is the most abrupt contributor in `r n_fen_abrupt` of the
  18 site and reconstruction combinations.
- The most important region does not depend on the AMOC pathway for
  `r n_robust` of the 18 site and reconstruction combinations.

# What the five numbers mean

The spider figures compare the source regions at one site with five numbers.
Each asks a simple question about the region's contribution to the δ¹⁸O
anomaly at that site. All five are computed on a common 500-yr grid between
21.5 and 10 ka. The example values are for NISA under GLAC-1D in the mixed
pathway.

**Share: how much of the signal comes from this region?** Average the
region's contribution over time, and divide by the sum of these averages over
all nine regions. The shares add up to 100 %. At NISA,
`r ex_top$region` has the largest share, `r pct(ex_top$share)` %.

**Dominant: how often is this region the biggest contributor?** Count the
500-yr steps in which the region contributes more than any other region, and
divide by the number of steps. A region can have a modest share but still
dominate for part of the time, for example during one large pulse. At NISA,
`r ex_dom$region` is the largest contributor in `r pct(ex_dom$dom)` % of the
steps.

**SD: how much does this region's contribution swing?** The standard
deviation of the region's contribution over time, in ‰. A region that
delivers a steady trickle has a small SD even if its share is large. At NISA,
`r ex_sd$region` swings the most (SD `r fmt(ex_sd$sd, 3)` ‰).

**Var. share: how much of the ups and downs of the total does this region
cause?** The covariance of the region's contribution with the total,
divided by the variance of the total, Cov(region, total) / Var(total). These
shares also add up to 100 %. A region that swings a lot, at the same times
as the total, gets a large value. A region that swings against the others can
get a negative value. At NISA, `r ex_vs$region` explains
`r pct(ex_vs$vshare)` % of the variance of the total.

**SD Δ: how abrupt is this region?** The standard deviation of the changes
from one 500-yr step to the next, in ‰. It is large when the region's
contribution jumps quickly, and small when it only drifts. At NISA,
`r ex_ab$region` is the most abrupt (SD Δ `r fmt(ex_ab$sd_diff, 3)` ‰).

In the spider figures, each of these numbers is one spoke. Each spoke is
scaled to the largest value of that number at the same site, so the outer
ring marks the site's maximum, and shapes can be compared between regions and
between the two reconstructions.

# Data and methods

**Forward model.** The source-region δ¹⁸O anomaly of each region (−35 ‰
end-member) is interpolated to 10-yr steps and convolved with the HadCM3
surface response to a 10-yr dye pulse in that region. The pulse responses
come from the runs xpran (cold), xprao (zonal) and xpujc (meridional). The
mixed pathway switches between them on the deglacial schedule. Water keeps
the response of the mode it was released in. A one-page schematic is at
https://claude.ai/artifact/CqfLtTKp2MCwDi9L8YFa6x.

**Proxy sites.** Ocean cores take the latitude-weighted mean over a ±2° box.
Speleothems and the ice core weight the surface field by the moisture uptake
of their precipitation, from the back-trajectory runs of Endres et al.
(2026b).

**Checks on the pulse responses.** The 50 decadal responses add up to the
500-yr constant-input equilibrium. At each site, the ratio of the summed
response to the equilibrium weights of Endres et al. (2026b) lies between
`r fmt(min(w_site$r))` and `r fmt(max(w_site$r))` for the cold and zonal
modes (median over all site, mode and region cells
`r fmt(median(chk$ratio, na.rm = TRUE))`). A constant forcing therefore
reaches the paper's equilibrium anomaly. As a direct test, each region's
source anomaly of the paper's four scenarios was held constant for 500 years
and run through the convolution. The resulting site anomalies agree with the
paper's (median ratio `r fmt(median(val$ratio_x1))`, 5–95 % range
`r fmt(quantile(val$ratio_x1, 0.05))`–`r fmt(quantile(val$ratio_x1, 0.95))`). Most of the
scatter is expected: the pulse responses come from the 17.8 ka simulations,
while the paper's 19.4 and 20.7 ka site values use those time slices' own
ocean states.
This is why the forcing gets no extra factor; the 2025.05 convolution
notebook multiplied it by 10, which gives `r fmt(median(val$ratio_x10), 1)`
times the paper's values. What the forward model changes is the timing:
short pulses arrive later and more spread out.

**Metrics.** The decadal site series were averaged into
`r length(unique(g$age))` steps of 500 years between 21.5 and 10 ka. This is
the rule of the previous report: GLAC-1D has 100-yr and ICE-6G 500-yr native
steps, and a common grid keeps resolution from looking like variability.

# Results

```{r}
#| label: fig-totals
#| fig-cap: "Total δ¹⁸O anomaly at the nine proxy sites, mixed AMOC pathway, decadal steps. Blue: GLAC-1D; orange: ICE-6G. The panels have their own vertical scales."
#| fig-width: 7
#| fig-height: 6
tot <- d |> filter(pathway == "mixed", time_bp <= win[1], time_bp >= win[2]) |>
  group_by(forcing, site, time_bp) |> summarise(anom = sum(anom), .groups = "drop") |>
  mutate(site = factor(site, levels = sites))
ggplot(tot, aes(time_bp / 1000, anom, colour = forcing)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~site, ncol = 3, scales = "free_y") +
  scale_x_reverse(breaks = seq(21, 10, -3)) +
  scale_colour_manual(values = rec_col, name = NULL) +
  labs(x = "Age (ka BP)", y = expression(delta^18*O ~ "anomaly at the site (‰)")) +
  theme_ms + theme(legend.position = "bottom")
```

@fig-totals shows the total anomaly at each site. @tbl-summary lists, for
each site, the region with the largest share, the largest SD and the largest
SD Δ in the mixed pathway. @tbl-pathways shows whether the most important
region changes with the AMOC pathway.

```{r}
#| label: tbl-summary
#| tbl-cap: "Mixed pathway, 21.5–10 ka, 500-yr grid. Most important: largest share (in brackets). Most variable: largest SD. Most abrupt: largest SD Δ."
mix |> mutate(site = factor(site, levels = sites)) |> arrange(site, forcing) |>
  transmute(Site = as.character(site), Forcing = forcing,
            `Most important` = sprintf("%s (%s %%)", important, pct(share)),
            `Most variable` = variable, `Most abrupt` = abrupt) |>
  mutate(Site = ifelse(duplicated(Site), "", Site)) |>
  knitr::kable()
```

```{r}
#| label: tbl-pathways
#| tbl-cap: "Most important source region per AMOC pathway. Codes: Med EIS_MedSea, Bri EIS_BayOfBiscay, Fen EIS_NorwegianSea, EurArc EIS_Arctic, AmeArc LAU_Arctic, GIS GIS_GreenlandSea, NLau LAU_LabradorSea, SLau LAU_StLawrence, GoM LAU_GulfOfMexico."
ss |> mutate(site = factor(site, levels = sites), col = paste(forcing, pathway),
             important = short[important]) |>
  select(site, col, important) |>
  pivot_wider(names_from = col, values_from = important) |>
  arrange(site) |>
  select(Site = site, `GLAC-1D cold`, `GLAC-1D zonal`, `GLAC-1D mixed`, `ICE-6G cold`, `ICE-6G zonal`, `ICE-6G mixed`) |>
  knitr::kable()
```

## Region profiles per site

The following figures show the five numbers for every source region at one
site, for GLAC-1D and ICE-6G in the mixed pathway. A large polygon means that
the region matters at that site. Where the blue and orange polygons differ,
the two reconstructions disagree about that region.

```{r}
#| label: fig-spider-nisa
#| fig-cap: "La Vallina (NISA), speleothem. Five numbers per source region (Share, Dominant, SD, Var. share, SD Δ), mixed pathway, 21.5–10 ka. Each spoke is scaled to the site's largest value of that number. Blue: GLAC-1D; orange: ICE-6G."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "La Vallina (NISA)")
```

```{r}
#| label: fig-spider-cwn
#| fig-cap: "Cave Without a Name, speleothem. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "Cave Without a Name")
```

```{r}
#| label: fig-spider-ngrip
#| fig-cap: "NGRIP, ice core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "NGRIP")
```

```{r}
#| label: fig-spider-md952010
#| fig-cap: "MD95-2010, Norwegian Sea core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "MD95-2010")
```

```{r}
#| label: fig-spider-ps2644
#| fig-cap: "PS2644-5, Denmark Strait core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "PS2644-5")
```

```{r}
#| label: fig-spider-md032664
#| fig-cap: "MD03-2664, Labrador Sea core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "MD03-2664")
```

```{r}
#| label: fig-spider-odp980
#| fig-cap: "ODP980, Feni Drift core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "ODP980")
```

```{r}
#| label: fig-spider-md952042
#| fig-cap: "MD95-2042, Iberian margin core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "MD95-2042")
```

```{r}
#| label: fig-spider-md022552
#| fig-cap: "MD02-2552, Gulf of Mexico core. As for NISA."
#| fig-width: 7
#| fig-height: 7
spider_plot(m, "MD02-2552")
```

# Caveats

- **Linearity.** The source anomaly is linear in discharge. Endres et al.
  (2026b) use a mixing fraction that saturates for large inputs, so the
  largest pulses are overstated here.
- **Fixed boundary conditions.** The cold and zonal pulse responses come from
  17.8 ka simulations. The boundary conditions of the meridional parent run
  (xpram) are not yet confirmed. The AMOC modes switch abruptly on a fixed
  schedule.
- **Meridional kernel.** The meridional pulse run branches from xpram, not
  from the paper's meridional run (xpral). It is used only in the mixed
  pathway between 14.7 and 13.7 ka.
- **Land sites.** NISA is weighted by the trajectory uptake masks. The 2025.05
  convolution notebook used a different moisture-source field, which gave a
  smaller NISA anomaly with the same timing.
- **Metrics.** The numbers describe the forward model, not the proxy
  records. Spokes are scaled per site, so a large polygon at one site cannot
  be compared in absolute terms with a large polygon at another.
